---
title: " Predicting Hotel Booking Outcomes Using Machine Learning"
author: "Petronela Tabalae, Divya Shori, Lorena Zovkic"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
    df_print: paged
---

```{r libraries setup, include = FALSE}
# Load all libraries at the start
library(tidyverse)   # core data science packages
library(readr)
library(httr)
library(jsonlite)
library(dplyr)
library(lubridate)
library(car)         # avPlot — added-variable plots
library(multcomp)    # glht — post-hoc contrasts
library(broom)       # tidy model output
library(kableExtra)  # enhanced tables
library(scales)      # axis formatting
library(patchwork)   # combining plots
library(caret)       # createDataPartition

# Get the hotel data
hotel_data <- read_csv(
  "H2.csv",
  na = c("NULL", "NA", ""),
  trim_ws = TRUE
)

# Fetch weather data
url <- "https://archive-api.open-meteo.com/v1/archive?latitude=38.72&longitude=-9.13&start_date=2015-01-01&end_date=2017-12-31&daily=temperature_2m_mean,precipitation_sum,sunshine_duration&timezone=Europe/Lisbon"
response <- GET(url)
weather_raw <- fromJSON(content(response, "text", encoding = "UTF-8"))
weather_daily <- as_tibble(weather_raw$daily) |>
  mutate(time = as.Date(time))

weather_monthly <- weather_daily |>
  mutate(
    year  = year(time),
    month = month.name[month(time)]
  ) |>
  group_by(year, month) |>
  summarise(
    avg_temp_c       = mean(temperature_2m_mean, na.rm = TRUE),
    precipitation_mm = sum(precipitation_sum,    na.rm = TRUE),
    sunshine_hours   = sum(sunshine_duration,    na.rm = TRUE) / 3600,
    .groups = "drop"
  )

# Join the hotel data
hotel_data <- hotel_data %>%
  left_join(weather_monthly,
            by = c("ArrivalDateYear" = "year",
                   "ArrivalDateMonth" = "month"))

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5)
```

```{r preprocessing, message=FALSE, warning=FALSE}
# ── FEATURE ENGINEERING ──────────────────────────────────────────────────────
hotel_data <- hotel_data %>%
  mutate(
    across(where(is.character), trimws),
    Agent   = ifelse(Agent   %in% c("NULL", ""), NA, Agent),
    Company = ifelse(Company %in% c("NULL", ""), NA, Company),
    total_stays = StaysInWeekendNights + StaysInWeekNights,
    arrival_season = case_when(
      ArrivalDateMonth %in% c("March", "April", "May")            ~ "Spring",
      ArrivalDateMonth %in% c("June", "July", "August")           ~ "Summer",
      ArrivalDateMonth %in% c("September", "October", "November") ~ "Autumn",
      TRUE                                                          ~ "Winter"
    ),
    is_via_agent = ifelse(is.na(Agent), 0L, 1L),
    country_region = case_when(
      Country %in% c("PRT","ESP","FRA","ITA","DEU","BEL",
                     "NLD","CHE","AUT","LUX","AND")               ~ "Western Europe",
      Country %in% c("GBR","IRL","DNK","SWE","NOR","FIN",
                     "ISL","POL","CZE","SVK","HUN")               ~ "N/E Europe",
      Country %in% c("USA","CAN","BRA","ARG","MEX","COL")         ~ "Americas",
      TRUE                                                          ~ "Other"
    ),
    Children = ifelse(is.na(Children), 0, Children)
  )

hotel_data <- hotel_data[, setdiff(names(hotel_data), 
                          c("ReservationStatus", "ReservationStatusDate", "Company"))]

hotel_data <- hotel_data %>%
  mutate(
    MarketSegment    = relevel(factor(MarketSegment),  ref = "Online TA"),
    CustomerType     = relevel(factor(CustomerType),   ref = "Transient"),
    DepositType      = relevel(factor(DepositType),    ref = "No Deposit"),
    Meal             = relevel(factor(Meal),           ref = "BB"),
    arrival_season   = relevel(factor(arrival_season), ref = "Summer"),
    country_region   = relevel(factor(country_region), ref = "Western Europe"),
    ReservedRoomType = factor(ReservedRoomType),
    IsCanceled       = as.integer(IsCanceled),
    IsRepeatedGuest  = as.integer(IsRepeatedGuest)
  )

cat("Dataset ready:", nrow(hotel_data), "bookings,", ncol(hotel_data), "variables\n")
mean_lt   <- round(mean(hotel_data$LeadTime), 1)
mean_temp <- round(mean(hotel_data$avg_temp_c, na.rm = TRUE), 1)
```

---

# Introduction

The hospitality industry operates under persistent uncertainty. Hotels must commit to staffing levels, room pricing, and inventory decisions well in advance, yet a significant share of bookings are cancelled before arrival. In city hotels particularly, where business and leisure travellers mix and booking patterns are highly seasonal, understanding what drives cancellations — and what shapes guest behaviour more broadly — has direct consequences for revenue management.

This report presents a machine learning analysis of booking records from a city hotel in Lisbon, Portugal. The data were collected and published by Antonio, Almeida, and Nunes (2019) and cover 79,330 individual reservations made between July 2015 and August 2017. Each record captures the full lifecycle of a booking: when it was made, who made it, what was requested, what was paid, and whether the guest ultimately arrived. To add environmental context, monthly weather observations for Lisbon — temperature, precipitation, and sunshine hours — are incorporated as complementary predictors, sourced from the Open-Meteo historical weather archive.

Three questions motivate the analysis. First, which booking and seasonal characteristics determine the price a guest pays for a room? Second, what kinds of bookings tend to result in longer stays? Third, what information available at the time of booking most reliably predicts whether a reservation will be cancelled?

To address these questions, six modelling approaches are applied: a Linear Model, a Poisson Generalised Linear Model, a Binomial Generalised Linear Model, a Generalised Additive Model, a Support Vector Machine, and a Neural Network. Together, these methods span the range from interpretable parametric models to flexible, data-driven approaches, allowing both explanation and prediction to be evaluated side by side. All models are trained on a held-out split of the data and evaluated on unseen test observations to ensure that reported performance reflects genuine out-of-sample accuracy.

---


 
# Exploratory Data Analysis
 
Understanding the data before modelling is not a preliminary step — it is the foundation of every decision that follows. This section examines what the dataset contains, what patterns emerge without any modelling assumptions, and what we found that required attention before the data could be used reliably. The findings here directly inform all preprocessing and modelling choices in the sections that follow.
 
---
 
## What Is in the Data?
 
The dataset covers 79,330 city hotel bookings from a single Lisbon property, recorded between July 2015 and August 2017. Each row represents one booking and carries information across four broad areas: who made the booking (guest type, origin, repeat status), how it was made (channel, agent, lead time), what was booked (room type, meal plan, special requests), and what happened (price paid, stay length, whether it was cancelled).
 
```{r eda-snapshot, echo=FALSE}
snapshot <- data.frame(
  Area = c("Who booked",
           "How it was booked",
           "What was booked",
           "What happened",
           "When they arrived",
           "Complementary weather"),
  Key_Variables = c(
    "CustomerType, Country, Adults, Children, Babies, IsRepeatedGuest",
    "MarketSegment, DistributionChannel, Agent, LeadTime, DepositType",
    "ReservedRoomType, Meal, TotalOfSpecialRequests, RequiredCarParkingSpaces",
    "IsCanceled, ADR, StaysInWeekendNights, StaysInWeekNights",
    "ArrivalDateYear, ArrivalDateMonth, arrival_season",
    "avg_temp_c, precipitation_mm, sunshine_hours"
  ),
  Notes = c(
    "74.9% Transient; 39% from Portugal",
    "48.9% Online TA; median lead time 74 days",
    "BB meal plan = 78.5% of bookings",
    "41.7% cancelled; median ADR €100",
    "2015–2017; August is busiest month",
    "Joined from Open-Meteo API on arrival month"
  )
)
 
kable(snapshot,
      caption = "Table 1: Dataset structure — 79,330 bookings, 31 variables across six areas.",
      col.names = c("Area", "Key Variables", "Notable Facts")) %>%
  kable_styling(full_width = TRUE, font_size = 11) %>%
  row_spec(0, bold = TRUE)
```
 
Three variables are excluded before any analysis begins. `ReservationStatus` and `ReservationStatusDate` directly encode the cancellation outcome and would constitute data leakage in any predictive model. `Company` is missing for 94.7% of bookings and carries no usable signal.
 
---
 
## How Are Bookings Distributed Over Time?
 
Demand is not uniform across the year. August is the busiest month by a considerable margin, and 2016 accounts for the largest share of bookings across the three years. This temporal structure has two implications: seasonal patterns are a genuine feature of the data, not noise, and any model that ignores seasonality is likely to underperform.
 
```{r eda-temporal, fig.cap="Figure EDA-1: Monthly booking volume across 2015–2017. August consistently peaks each year, while January and December are the quietest months. The seasonal rhythm is stable across all three years, confirming that arrival season is a reliable and meaningful predictor for all models."}
 
hotel_data %>%
  mutate(
    ArrivalDateMonth = factor(ArrivalDateMonth, levels = month.name),
    Year = factor(ArrivalDateYear)
  ) %>%
  count(ArrivalDateMonth, Year) %>%
  ggplot(aes(x = ArrivalDateMonth, y = n, colour = Year, group = Year)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("2015" = "#2166ac",
                                  "2016" = "#4dac26",
                                  "2017" = "#d01c8b")) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(
    title    = "Monthly Booking Volume by Year",
    subtitle = "Consistent seasonal pattern across all three years",
    x = NULL, y = "Number of Bookings", colour = "Year"
  ) +
  theme_minimal(base_size = 11)
```
 
---
 
## What Does the Typical Booking Look Like?
 
Before exploring variation, it helps to establish what a typical booking in this dataset looks like. The median guest books 74 days in advance, stays 3 nights, pays €100 per night, and does not cancel. That is the baseline. Everything the models do is an attempt to explain departures from that baseline.
 
```{r eda-distributions, fig.cap="Figure EDA-2: Distributions of the four core numeric variables. ADR and lead time are right-skewed with long upper tails, indicating that log-transformation or careful scaling will be needed before modelling. Total nights and special requests are discrete counts concentrated at low values, consistent with count distributions.", fig.height=5}
 
p1 <- ggplot(hotel_data %>% filter(ADR > 0, ADR < 500), aes(x = ADR)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = median(hotel_data$ADR[hotel_data$ADR > 0]),
             linetype = "dashed", colour = "darkred", linewidth = 0.8) +
  annotate("text", x = median(hotel_data$ADR[hotel_data$ADR > 0]) + 20,
           y = Inf, vjust = 2,
           label = paste0("Median €", median(hotel_data$ADR[hotel_data$ADR > 0])),
           size = 3, colour = "darkred") +
  scale_x_continuous(labels = dollar_format(prefix = "€")) +
  labs(title = "Room Price (ADR)", x = "€/night", y = "Count") +
  theme_minimal(base_size = 10)
 
p2 <- ggplot(hotel_data, aes(x = LeadTime)) +
  geom_histogram(bins = 60, fill = "darkorange", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = median(hotel_data$LeadTime),
             linetype = "dashed", colour = "darkred", linewidth = 0.8) +
  annotate("text", x = median(hotel_data$LeadTime) + 20,
           y = Inf, vjust = 2,
           label = paste0("Median ", median(hotel_data$LeadTime), " days"),
           size = 3, colour = "darkred") +
  labs(title = "Lead Time", x = "Days before arrival", y = "Count") +
  theme_minimal(base_size = 10)
 
p3 <- hotel_data %>%
  filter(total_stays > 0, total_stays <= 12) %>%
  count(total_stays) %>%
  ggplot(aes(x = total_stays, y = n)) +
  geom_col(fill = "seagreen", colour = "white", alpha = 0.85) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Total Nights Booked", x = "Nights", y = "Count") +
  theme_minimal(base_size = 10)
 
p4 <- hotel_data %>%
  count(TotalOfSpecialRequests) %>%
  ggplot(aes(x = TotalOfSpecialRequests, y = n)) +
  geom_col(fill = "mediumpurple", colour = "white", alpha = 0.85) +
  scale_x_continuous(breaks = 0:5) +
  labs(title = "Special Requests", x = "Number of requests", y = "Count") +
  theme_minimal(base_size = 10)
 
(p1 + p2) / (p3 + p4)
```
 
---
 
## Where Does the Variation Come From?
 
The most important question in EDA is not what the average looks like — it is what drives variation. We found that market segment is by far the most powerful organising variable in this dataset. It separates bookings by price, stay length, and cancellation rate simultaneously.
 
```{r eda-segment-overview, fig.cap="Figure EDA-3: Mean ADR (€), mean stay length (nights), and cancellation rate by market segment. Online TA dominates volume at 48.9% of bookings. Direct bookings command the highest average price. Corporate guests stay the shortest. Groups cancel at nearly 69%. These three dimensions — price, stay length, cancellation — define the three core research questions of this project."}
 
seg_summary <- hotel_data %>%
  filter(!as.character(MarketSegment) %in% c("Undefined")) %>%
  group_by(MarketSegment) %>%
  summarise(
    mean_adr       = mean(ADR[ADR > 0], na.rm = TRUE),
    mean_stays     = mean(total_stays[total_stays > 0], na.rm = TRUE),
    cancel_rate    = mean(IsCanceled),
    n              = n(),
    .groups = "drop"
  ) %>%
  mutate(share = n / sum(n))
 
p_adr_seg <- ggplot(seg_summary,
                    aes(x = reorder(as.character(MarketSegment), mean_adr),
                        y = mean_adr, fill = as.character(MarketSegment))) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = paste0("€", round(mean_adr))),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 145)) +
  labs(title = "Mean ADR", x = NULL, y = "€/night") +
  theme_minimal(base_size = 10)
 
p_stays_seg <- ggplot(seg_summary,
                      aes(x = reorder(as.character(MarketSegment), mean_adr),
                          y = mean_stays, fill = as.character(MarketSegment))) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = round(mean_stays, 1)),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 5)) +
  labs(title = "Mean Stay Length", x = NULL, y = "Nights") +
  theme_minimal(base_size = 10)
 
p_cancel_seg <- ggplot(seg_summary,
                       aes(x = reorder(as.character(MarketSegment), mean_adr),
                           y = cancel_rate, fill = as.character(MarketSegment))) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(cancel_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 0.95)) +
  labs(title = "Cancellation Rate", x = NULL, y = NULL) +
  theme_minimal(base_size = 10)
 
p_adr_seg + p_stays_seg + p_cancel_seg
```
 
---
 
## What Data Quality Issues Did We Find?
 
Three issues required specific attention and shaped how the data was prepared for modelling.
 
**Issue 1 — Zero ADR records.** 1,208 bookings (1.5%) have an ADR of zero. These are complimentary rooms — provided free of charge — and are not reflective of market pricing. Including them would distort any model of room price. They are excluded from the Linear Model but retained in classification models where price is a predictor rather than the outcome.
 
**Issue 2 — The Non Refund deposit paradox.** Bookings with a Non Refund deposit show a 99.8% cancellation rate across 12,868 records. This is almost certainly not genuine guest behaviour — it likely reflects a recording convention where no-shows or agent-initiated cancellations are logged as cancelled regardless of whether payment was collected. This variable is included in the classification models but flagged explicitly; its coefficient should not be interpreted causally.
 
**Issue 3 — Skewness in key variables.** ADR and lead time are both strongly right-skewed. ADR is log-transformed for the Linear Model. Lead time and temperature are mean-centred before modelling so that model intercepts represent the most common booking profile rather than an impossible baseline of zero days and zero degrees.
 
```{r eda-quality, fig.cap="Figure EDA-4: Cancellation rate by deposit type (left) and ADR distribution highlighting zero-price bookings (right). The near-100% cancellation rate for Non Refund deposits and the cluster of zero-ADR bookings are the two most important data quality findings — both are addressed in data preparation before any model is fitted."}
 
p_dep <- hotel_data %>%
  group_by(DepositType) %>%
  summarise(
    rate = mean(IsCanceled),
    n    = n(),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(DepositType, rate), y = rate, fill = DepositType)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = paste0(scales::percent(rate, accuracy = 0.1),
                               "\n(n=", scales::comma(n), ")")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("No Deposit"  = "steelblue",
                               "Non Refund"  = "tomato",
                               "Refundable"  = "darkorange")) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 1.2)) +
  labs(title = "Cancellation Rate by Deposit Type",
       subtitle = "Non Refund = 99.8% — likely a recording artefact",
       x = NULL, y = "Cancellation Rate") +
  theme_minimal(base_size = 10)
 
p_zero_adr <- hotel_data %>%
  mutate(zero_adr = ifelse(ADR == 0, "Zero ADR\n(complimentary)", "ADR > 0")) %>%
  filter(ADR < 500) %>%
  ggplot(aes(x = ADR, fill = zero_adr)) +
  geom_histogram(bins = 70, colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("Zero ADR\n(complimentary)" = "tomato",
                               "ADR > 0" = "steelblue"),
                    name = NULL) +
  scale_x_continuous(labels = dollar_format(prefix = "€")) +
  labs(title = "ADR Distribution",
       subtitle = "1,208 zero-ADR bookings excluded from Linear Model",
       x = "ADR (€/night)", y = "Count") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
 
p_dep + p_zero_adr
```
 
---
 
## How Does Cancellation Vary With Lead Time?
 
One relationship in the data stands out as particularly striking and has direct implications for how models are specified. The probability of cancellation is not a straight line — it rises sharply with lead time up to around 150 days, then begins to plateau. A guest who books 200 days in advance is far more likely to cancel than one who books a week out.
 
```{r eda-leadtime, fig.cap="Figure EDA-5: Estimated cancellation probability as a function of lead time, fitted with a LOESS smoother. The relationship is clearly nonlinear — risk rises steeply for bookings made more than 30 days in advance before levelling off. A straight line would substantially misrepresent this pattern."}
 
hotel_data %>%
  ggplot(aes(x = LeadTime, y = IsCanceled)) +
  geom_smooth(method = "loess", span = 0.35,
              colour = "steelblue", fill = "lightblue",
              linewidth = 1.2) +
  geom_hline(yintercept = mean(hotel_data$IsCanceled),
             linetype = "dashed", colour = "darkred", linewidth = 0.7) +
  annotate("text", x = 420,
           y = mean(hotel_data$IsCanceled) + 0.03,
           label = paste0("Overall rate = ",
                          scales::percent(mean(hotel_data$IsCanceled),
                                          accuracy = 0.1)),
           colour = "darkred", size = 3.2) +
  scale_x_continuous(limits = c(0, 500)) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 0.85)) +
  labs(
    title    = "Cancellation Probability by Lead Time",
    subtitle = "LOESS smoother — the nonlinear shape is a genuine feature of the data",
    x = "Lead Time (days before arrival)",
    y = "Estimated Cancellation Probability"
  ) +
  theme_minimal(base_size = 11)
```
 
---
 
## What This Means for Data Preparation
 
The EDA surfaces five decisions that carry through to every model in this report:
 
**1. Log-transform ADR.** The right-skewed distribution of room price requires a log-transformation before it can serve as a response variable in the Linear Model.
 
**2. Remove zero-ADR records from pricing models.** Complimentary bookings are not priced by the market and would distort any model predicting room price.
 
**3. Remove same-day bookings from stay-length models.** 331 bookings with zero total nights are excluded from the Poisson model as they represent no meaningful stay.
 
**4. Centre continuous predictors.** Lead time and temperature are mean-centred before modelling so intercepts are interpretable at the most common booking profile.
 
**5. Treat Non Refund deposit with caution.** Its near-perfect association with cancellation reflects a data recording issue, not a causal relationship. It is included as a predictor but not interpreted causally.
 
These five decisions are applied in the preprocessing step before any model is fitted and remain consistent across all six models that follow.
 
---

# Model 1: Linear Model — What Drives Room Price?

*Section lead: Divya Shori*

## Research Question

> **Can we predict the room price for an incoming booking, and which factors drive the most meaningful price variation?**

Room price measured as Average Daily Rate (ADR, euros per night), is the primary lever through which hotels manage revenue. Understanding what drives ADR matters because even small systematic differences in pricing across segments or seasons can translate into significant revenue at scale. This model asks: given everything we know about a booking at the moment it is made, what room price should we expect?

## Why This Model

The Linear Model is the appropriate starting point for a continuous numeric response variable. ADR is a real-valued amount, and once log-transformed to address skewness, the assumptions of normality and constant variance are reasonably satisfied. The Linear Model also produces directly interpretable coefficients, each one tells us the percentage change in room price associated with a one-unit change in a predictor, holding everything else constant. This transparency is a significant advantage over more complex models.

## Data Preparation

Two subsets of the data are excluded before modelling. First, 1,208 bookings with an ADR of zero are removed, these represent complimentary rooms and do not reflect market pricing. Second, the Complementary and Undefined market segments are excluded for the same reason. The remaining 78,120 bookings form the modelling dataset.

The response variable ADR is log-transformed. The raw distribution is strongly right-skewed, a mean of €105 with a tail extending to €5,400, which would violate the normality assumption of the Linear Model. Log-transformation produces a near-symmetric distribution and makes coefficients interpretable as multiplicative percentage effects on price, which is more intuitive for a business audience than raw euro-unit effects.

Two continuous predictors are mean-centred before fitting: lead time (mean = `r mean_lt` days) and monthly temperature (mean = `r mean_temp`°C). Centring ensures the model intercept has a meaningful interpretation — the expected room price for the most common booking profile at average lead time and average temperature — rather than an extrapolation to impossible values like zero days or zero degrees.

```{r lm-motivation-plot, fig.cap="Figure 1: Raw ADR (left) is strongly right-skewed with a long upper tail, violating the normality assumption of the Linear Model. After log-transformation (right), the distribution is approximately symmetric and unimodal — confirming that log(ADR) is the appropriate response variable."}

hotel_lm <- hotel_data %>%
  filter(ADR > 0,
         !as.character(MarketSegment) %in% c("Complementary", "Undefined")) %>%
  mutate(
    log_adr      = log(ADR),
    LeadTime_c   = LeadTime   - mean(LeadTime,   na.rm = TRUE),
    avg_temp_c_c = avg_temp_c - mean(avg_temp_c, na.rm = TRUE)
  )
hotel_lm$MarketSegment <- droplevels(hotel_lm$MarketSegment)

mean_lt   <- round(mean(hotel_data$LeadTime), 1)
mean_temp <- round(mean(hotel_data$avg_temp_c, na.rm = TRUE), 1)

p_raw <- ggplot(hotel_lm, aes(x = ADR)) +
  geom_histogram(bins = 80, fill = "tomato", colour = "white", alpha = 0.85) +
  scale_x_continuous(labels = dollar_format(prefix = "€"), limits = c(0, 600)) +
  labs(title = "Raw ADR", subtitle = "Right-skewed — not suitable for LM",
       x = "ADR (€/night)", y = "Count") +
  theme_minimal(base_size = 11)

p_log <- ggplot(hotel_lm, aes(x = log_adr)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(hotel_lm$log_adr),
             linetype = "dashed", colour = "black", linewidth = 0.8) +
  annotate("text", x = mean(hotel_lm$log_adr) + 0.15, y = Inf, vjust = 2,
           label = paste0("Mean = €", round(exp(mean(hotel_lm$log_adr)), 1)),
           size = 3.5) +
  labs(title = "log(ADR)", subtitle = "Symmetric — appropriate for LM",
       x = "log(ADR)", y = "Count") +
  theme_minimal(base_size = 11)

p_raw + p_log
```

## Predictor Exploration

Before fitting the model, we examine how key predictors relate to log(ADR). This motivates the predictor set and informs decisions about reference level choice and centering.

```{r lm-eda, fig.cap="Figure 2: log(ADR) by market segment (left) — Direct and Online TA channels command the highest median rates, while Corporate and Groups anchor the lower end. The relationship between lead time and log(ADR) (right) reveals a negative association: bookings made further in advance tend to secure lower prices, consistent with early-bird discount pricing strategies."}

p_seg <- ggplot(hotel_lm,
                aes(x = reorder(MarketSegment, log_adr, median),
                    y = log_adr, fill = MarketSegment)) +
  geom_violin(trim = TRUE, alpha = 0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA,
               show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0("€", round(exp(x))),
    sec.axis = sec_axis(~ ., name = "log(ADR)")
  ) +
  labs(title = "Room Price by Market Segment",
       x = NULL, y = "ADR (€) [log scale]") +
  theme_minimal(base_size = 10)

p_lead <- ggplot(hotel_lm %>% sample_n(8000),
                 aes(x = LeadTime, y = log_adr)) +
  geom_point(alpha = 0.12, size = 0.7, colour = "steelblue") +
  geom_smooth(method = "loess", span = 0.4,
              colour = "darkred", fill = "salmon", linewidth = 1.1) +
  geom_vline(xintercept = mean_lt, linetype = "dotted",
             colour = "grey40", linewidth = 0.8) +
  annotate("text", x = mean_lt + 15, y = max(hotel_lm$log_adr) * 0.98,
           label = paste0("Mean = ", mean_lt, " days"), size = 3, colour = "grey40") +
  scale_x_continuous(limits = c(0, 400)) +
  labs(title = "Lead Time vs. Room Price",
       x = "Lead Time (days before arrival)", y = "log(ADR)") +
  theme_minimal(base_size = 10)

p_seg + p_lead
```

## Model Setup: Centering and Reference Levels

Two methodological choices follow directly from the lecture material. First, continuous predictors (`LeadTime`, `avg_temp_c`) are **mean-centred**: subtracting the sample mean (`r mean_lt` days for lead time; `r mean_temp`°C for temperature) ensures that the model intercept has a concrete interpretation — the expected log(ADR) for a booking made with average lead time at average temperature, in the Online TA reference segment, during Summer. Without centering, the intercept would represent a booking made 0 days in advance at 0°C — a combination that does not exist in the data.

Second, `Online TA` is set as the reference level for market segment (the largest group at 49% of bookings), and `Transient` for customer type (75% of bookings). This choice ensures the intercept represents the most common booking profile, making the reference case immediately meaningful.

## Model Fitting

```{r lm-split}
set.seed(2026)
idx_lm   <- createDataPartition(hotel_lm$log_adr, p = 0.8, list = FALSE)
train_lm <- hotel_lm[idx_lm, ]
test_lm  <- hotel_lm[-idx_lm, ]
cat("Training:", nrow(train_lm), "| Test:", nrow(test_lm), "\n")
```

```{r lm-fit}
model_lm <- lm(
  log_adr ~ arrival_season + MarketSegment + CustomerType +
            ReservedRoomType + LeadTime_c + TotalOfSpecialRequests +
            IsRepeatedGuest + Meal + is_via_agent + avg_temp_c_c,
  data = train_lm
)

cat(sprintf("Adjusted R²  : %.3f\n", summary(model_lm)$adj.r.squared))
cat(sprintf("Residual SE  : %.3f log-ADR units\n", summary(model_lm)$sigma))
```

## Formal Predictor Testing with drop1()

Rather than relying solely on individual coefficient p-values, we apply `drop1()` with an F-test to assess the contribution of each predictor to the overall model. This tests whether removing a predictor significantly worsens model fit — a more honest measure of importance than a coefficient's t-statistic, particularly for multi-level factors where no single coefficient captures the full effect of the variable.

```{r lm-drop1, fig.cap="Figure 3: F-statistics from drop1() for each predictor in the Linear Model. The red dashed line marks the critical F-value at α = 0.05. All predictors cross the threshold, confirming that none are redundant — every variable in the model carries statistically significant explanatory power over room price."}

drop1_res <- drop1(model_lm, test = "F")

drop1_df <- as.data.frame(drop1_res) %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "<none>") %>%
  mutate(Predictor = dplyr::recode(Predictor,
    "arrival_season"         = "Arrival Season",
    "MarketSegment"          = "Market Segment",
    "CustomerType"           = "Customer Type",
    "ReservedRoomType"       = "Reserved Room Type",
    "LeadTime_c"             = "Lead Time (centred)",
    "TotalOfSpecialRequests" = "Special Requests",
    "IsRepeatedGuest"        = "Repeated Guest",
    "Meal"                   = "Meal Plan",
    "is_via_agent"           = "Booked via Agent",
    "avg_temp_c_c"           = "Avg Temperature (centred)"
  ))

crit_f <- qf(0.95, df1 = 1, df2 = df.residual(model_lm))

ggplot(drop1_df, aes(x = `F value`, y = reorder(Predictor, `F value`))) +
  geom_segment(aes(x = 0, xend = `F value`,
                   yend = reorder(Predictor, `F value`)),
               colour = "grey75", linewidth = 0.8) +
  geom_point(aes(colour = `Pr(>F)` < 0.05), size = 4) +
  scale_colour_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey50"),
    labels = c("TRUE" = "Significant (p < 0.05)", "FALSE" = "Not significant"),
    name   = NULL
  ) +
  geom_vline(xintercept = crit_f, linetype = "dashed",
             colour = "darkred", linewidth = 0.8) +
  annotate("text", x = crit_f + 5, y = 1,
           label = paste0("Critical F = ", round(crit_f, 1)),
           colour = "darkred", size = 3.2, hjust = 0) +
  labs(
    title    = "Predictor Importance: drop1() F-Statistics",
    subtitle = "Each bar shows how much model fit deteriorates when that predictor is removed",
    x = "F-statistic", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## Key Results

The `drop1()` test confirms that every predictor in the model contributes significantly, none can be removed without meaningfully worsening fit. The strongest predictors are market segment, room type, and arrival season, consistent with the exploratory analysis. Lead time and temperature, while smaller in absolute effect, are both statistically robust and carry practical significance.

Three findings stand out for interpretation. First, market segment is the single largest 
driver of price variation: Direct bookings command the highest rates, while Corporate and 
Group bookings anchor the lower end — reflecting the commission pressure and volume 
discounts associated with intermediary and group channels respectively. Second, Summer 
arrivals attract a meaningful price premium over other seasons, consistent with peak leisure demand in Lisbon. Third, each additional day of lead time is associated with a small but consistent decrease in room price, suggesting that guests who book early secure better rates, a pattern the hotel could leverage through targeted early-booking promotions.

## The Unique Effect of Lead Time: Added-Variable Plot

A coefficient in a multiple regression reflects the effect of a predictor *after accounting for all other predictors*. To visualise this unambiguously for lead time — the most actionable predictor for pricing strategy — we construct an **added-variable (partial regression) plot**. This plot shows the residual variation in log(ADR) that cannot be explained by any other predictor in the model, plotted against the residual variation in lead time after removing the same set of predictors. The slope of the relationship is therefore the pure, isolated effect of lead time on price.

```{r lm-avplot, fig.cap="Figure 4: Added-variable plot for lead time. The negative slope confirms that longer lead times are genuinely associated with lower room prices — this is not an artefact of confounding with season or market segment."}

model_no_lt <- lm(
  log_adr ~ arrival_season + MarketSegment + CustomerType +
            ReservedRoomType + TotalOfSpecialRequests +
            IsRepeatedGuest + Meal + is_via_agent + avg_temp_c_c,
  data = train_lm
)

model_lt_on_x <- lm(
  LeadTime_c ~ arrival_season + MarketSegment + CustomerType +
               ReservedRoomType + TotalOfSpecialRequests +
               IsRepeatedGuest + Meal + is_via_agent + avg_temp_c_c,
  data = train_lm
)

av_df <- data.frame(
  resid_y  = residuals(model_no_lt),
  resid_lt = residuals(model_lt_on_x)
)

ggplot(av_df %>% sample_n(6000), aes(x = resid_lt, y = resid_y)) +
  geom_point(alpha = 0.12, size = 0.7, colour = "steelblue") +
  geom_smooth(method = "lm", colour = "darkred",
              fill = "salmon", linewidth = 1.2, se = TRUE) +
  labs(
    title    = "Added-Variable Plot: Lead Time → Room Price",
    subtitle = "Pure effect of lead time after controlling for all other predictors",
    x        = "Lead Time residuals (days — unexplained by other predictors)",
    y        = "log(ADR) residuals (unexplained by other predictors)"
  ) +
  theme_minimal(base_size = 11)
```

## Marginal Effect: Translating Lead Time into Euros

```{r lm-marginal, fig.cap="Figure 5: Predicted ADR (€) as a function of lead time for a typical Online TA Transient Summer booking. A last-minute booking commands roughly €20 more per night than one made 200 days in advance."}

lead_seq    <- seq(0, 400, by = 2)
mean_lt_val <- mean(train_lm$LeadTime, na.rm = TRUE)
mode_room   <- names(sort(table(train_lm$ReservedRoomType), decreasing = TRUE))[1]

newdata_lt <- data.frame(
  LeadTime_c             = lead_seq - mean_lt_val,
  MarketSegment          = factor("Online TA",  levels = levels(train_lm$MarketSegment)),
  CustomerType           = factor("Transient",  levels = levels(train_lm$CustomerType)),
  arrival_season         = factor("Summer",     levels = levels(train_lm$arrival_season)),
  ReservedRoomType       = factor(mode_room,    levels = levels(train_lm$ReservedRoomType)),
  Meal                   = factor("BB",         levels = levels(train_lm$Meal)),
  TotalOfSpecialRequests = mean(train_lm$TotalOfSpecialRequests, na.rm = TRUE),
  IsRepeatedGuest        = 0L,
  is_via_agent           = 1L,
  avg_temp_c_c           = 0
)

preds_lt       <- predict(model_lm, newdata = newdata_lt, interval = "confidence")
newdata_lt$fit <- exp(preds_lt[, "fit"])
newdata_lt$lwr <- exp(preds_lt[, "lwr"])
newdata_lt$upr <- exp(preds_lt[, "upr"])

ggplot(newdata_lt, aes(x = lead_seq)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = fit), colour = "steelblue", linewidth = 1.3) +
  geom_vline(xintercept = mean_lt_val, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  annotate("text", x = mean_lt_val + 8, y = max(newdata_lt$upr) * 0.98,
           label = paste0("Mean lead time\n(", round(mean_lt_val), " days)"),
           size = 3.2, colour = "grey40", hjust = 0) +
  scale_y_continuous(labels = dollar_format(prefix = "€")) +
  labs(
    title    = "Marginal Effect of Lead Time on Room Price",
    subtitle = "Reference: Online TA · Transient · Summer · BB meal · via agent",
    x        = "Lead Time (days before arrival)",
    y        = "Predicted ADR (€/night)"
  ) +
  theme_minimal(base_size = 11)
```

## Model Diagnostics

```{r lm-diagnostics, fig.cap="Figure 6: Residual diagnostics — broadly homoscedastic spread and approximate normality support the adequacy of the log-transformation.", fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(model_lm, which = 1, main = "Residuals vs Fitted",
     col = adjustcolor("steelblue", 0.3), pch = 16, cex = 0.5)
plot(model_lm, which = 2, main = "Normal Q-Q",
     col = adjustcolor("steelblue", 0.5), pch = 16, cex = 0.5)
par(mfrow = c(1, 1))
```

## Test Set Performance

```{r lm-test}
pred_lm  <- predict(model_lm, newdata = test_lm)
rmse_lm  <- sqrt(mean((pred_lm - test_lm$log_adr)^2))
r2_test  <- cor(pred_lm, test_lm$log_adr)^2
rmse_eur <- exp(mean(test_lm$log_adr) + rmse_lm) - exp(mean(test_lm$log_adr))

cat(sprintf("Adjusted R² (train): %.3f\n", summary(model_lm)$adj.r.squared))
cat(sprintf("R² on test set:      %.3f\n", r2_test))
cat(sprintf("Approx. error in €:  ±€%.1f per night\n", abs(rmse_eur)))
```

```{r lm-pred-vs-actual, fig.cap="Figure 7: Predicted vs. observed ADR on the held-out test set. Points cluster tightly around the 45° line, confirming good out-of-sample generalisation."}

data.frame(predicted = pred_lm, observed = test_lm$log_adr) %>%
  sample_n(3000) %>%
  ggplot(aes(x = exp(predicted), y = exp(observed))) +
  geom_point(alpha = 0.2, size = 0.8, colour = "steelblue") +
  geom_abline(slope = 1, intercept = 0,
              colour = "darkred", linetype = "dashed", linewidth = 0.9) +
  scale_x_continuous(labels = dollar_format(prefix = "€"), limits = c(0, 600)) +
  scale_y_continuous(labels = dollar_format(prefix = "€"), limits = c(0, 600)) +
  labs(
    title    = "Predicted vs. Observed ADR — Test Set",
    subtitle = "Red dashed line = perfect prediction",
    x = "Predicted ADR (€/night)", y = "Observed ADR (€/night)"
  ) +
  theme_minimal(base_size = 11)
```

## Business Takeaway

The Linear Model explains `r round(summary(model_lm)$adj.r.squared * 100, 1)`% of variance in log(ADR) on the training set, with an out-of-sample R² of `r round(r2_test, 3)` and a practical prediction error of approximately ±€`r round(abs(rmse_eur), 1)` per night.  The model explains a substantial proportion of room price variance and predicts unseen bookings to within approximately ±€`r round(abs(rmse_eur), 1)` per night. For a revenue manager, this has two direct applications. First, the model can serve as a real-time pricing benchmark: any incoming booking can be scored against the model's prediction, and bookings where the proposed rate deviates significantly from the prediction can be flagged for manual review. Second, the segment and season coefficients provide a clear, data-driven basis for setting differential pricing strategies — knowing that Direct bookings sustain higher rates than Online TA, for instance, justifies investing in direct channel marketing to shift the booking mix toward higher-margin customers.


---

# Model 2: Poisson GLM — Who Stays Longest?

*Section lead: Divya Shori*

## Research Question

> **Which guest segments and booking characteristics predict longer stays, and can we quantify the premium value of high-stay-length segments?**

Length of stay is one of the most underused levers in hotel revenue management. A guest who stays four nights generates roughly twice the revenue of a guest who stays two, with lower relative operational cost per night. Yet marketing and channel strategies are rarely optimised for stay length, most focus on cancellation or price. This model asks: which types of guests book the longest stays, and can we identify the segments most worth targeting for long-stay promotions?

## Why This Model

Total nights booked (`total_stays`) is a non-negative integer count variable. A standard 
Linear Model is inappropriate for count data — it can predict negative values and does not respect the discrete, bounded nature of the response. The Poisson Generalised Linear Model is designed for exactly this type of outcome: it models the log of the expected count as a linear function of predictors, guaranteeing non-negative predictions and producing coefficients interpretable as multiplicative rate ratios. Before accepting the Poisson assumption however, we formally test whether the data are overdispersed — a common violation in real-world count data that requires a correction.

## Data Preparation

Bookings with zero total nights (n = 331, likely same-day cancellations or data entry 
errors) are excluded from the analysis, as a zero-night stay cannot be meaningfully 
modelled as a count outcome. The two remaining observations in the Undefined market segment are also excluded. All other preprocessing — factor relevelling, feature engineering, weather join, was applied in the shared preprocessing step above.

## Count Distribution and Overdispersion Motivation

```{r poisson-dist, fig.cap="Figure 8: Observed stay length (blue bars) versus the theoretical Poisson distribution at the observed mean (red line). The heavier right tail in the data relative to the Poisson prediction motivates a formal overdispersion check."}

hotel_pois <- hotel_data %>%
  filter(total_stays > 0,
         !as.character(MarketSegment) %in% c("Undefined"))
hotel_pois$MarketSegment <- droplevels(hotel_pois$MarketSegment)

obs_lambda  <- mean(hotel_pois$total_stays)
max_display <- 14

pois_theoretical <- data.frame(
  nights   = 1:max_display,
  expected = dpois(1:max_display, lambda = obs_lambda) * nrow(hotel_pois)
)

hotel_pois %>%
  filter(total_stays <= max_display) %>%
  ggplot(aes(x = total_stays)) +
  geom_histogram(aes(y = after_stat(count)), binwidth = 1,
                 fill = "steelblue", colour = "white", alpha = 0.75) +
  geom_line(data = pois_theoretical,
            aes(x = nights, y = expected),
            colour = "darkred", linewidth = 1.3) +
  geom_point(data = pois_theoretical,
             aes(x = nights, y = expected),
             colour = "darkred", size = 2.5) +
  scale_x_continuous(breaks = 1:max_display) +
  labs(
    title    = "Distribution of Total Nights Booked",
    subtitle = paste0("Observed (blue) vs. Poisson(λ = ", round(obs_lambda, 2),
                      ") prediction (red)"),
    x = "Total Nights Booked", y = "Number of Bookings"
  ) +
  theme_minimal(base_size = 11)
```

```{r poisson-meanvar, fig.cap="Figure 9: Mean versus variance of total nights booked across market segment × season groups. Nearly all points lie above the Var = Mean line, confirming systematic overdispersion — variance exceeds the mean in almost every subgroup."}

mv_data <- hotel_pois %>%
  group_by(MarketSegment, arrival_season) %>%
  summarise(
    group_mean = mean(total_stays),
    group_var  = var(total_stays),
    n          = n(),
    .groups    = "drop"
  ) %>%
  filter(n >= 30)

ggplot(mv_data, aes(x = group_mean, y = group_var,
                    colour = MarketSegment, size = n)) +
  geom_point(alpha = 0.85) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "darkred", linewidth = 0.9) +
  annotate("text", x = 1.3, y = 1.2,
           label = "Var = Mean\n(Poisson assumption)",
           colour = "darkred", size = 3.2, hjust = 0) +
  scale_size_continuous(range = c(2, 9), name = "Bookings in group") +
  scale_colour_brewer(palette = "Set2", name = "Market Segment") +
  labs(
    title    = "Mean–Variance Relationship by Market Segment and Season",
    subtitle = "Points above the dashed line signal overdispersion (Var > Mean)",
    x = "Group Mean (nights)", y = "Group Variance"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

## Model Fitting and Overdispersion Check

```{r poisson-fit}
set.seed(2026)
idx_pois   <- createDataPartition(hotel_pois$total_stays, p = 0.8, list = FALSE)
train_pois <- hotel_pois[idx_pois, ]
test_pois  <- hotel_pois[-idx_pois, ]

model_pois <- glm(
  total_stays ~ MarketSegment + CustomerType + arrival_season +
                DepositType + IsRepeatedGuest + Meal +
                avg_temp_c + precipitation_mm + country_region,
  data = train_pois, family = poisson
)

disp_ratio <- deviance(model_pois) / df.residual(model_pois)
cat(sprintf("Dispersion ratio (Deviance / df): %.2f\n", disp_ratio))
cat(sprintf("→ %s\n",
    ifelse(disp_ratio > 1.5,
           "Overdispersion confirmed — refitting with quasi-Poisson.",
           "No strong overdispersion — Poisson retained.")))
```

```{r quasipoisson-refit}
if (disp_ratio > 1.5) {
  model_pois <- glm(
    total_stays ~ MarketSegment + CustomerType + arrival_season +
                  DepositType + IsRepeatedGuest + Meal +
                  avg_temp_c + precipitation_mm + country_region,
    data = train_pois, family = quasipoisson
  )
  cat("Quasi-Poisson fitted. Coefficients unchanged; SEs corrected upward.\n")
}

model_null <- glm(total_stays ~ 1, data = train_pois, family = quasipoisson)
anova(model_null, model_pois, test = "F")
```

## Formal Predictor Testing

```{r poisson-drop1}
drop1(model_pois, test = "F")
```

## Key Results

The `drop1()` F-tests confirm that market segment, customer type, meal plan, and seasonal conditions all contribute significantly to stay length. The rate ratio forest plot 
(Figure 10) reveals the direction and magnitude of these effects after controlling for 
confounders. Corporate and Contract customers book the shortest stays — consistent with 
single-night business travel — while Aviation and Online TA guests tend toward longer stays. Summer arrivals book longer stays than other seasons, and guests on half-board or full-board meal plans stay notably longer than bed-and-breakfast guests, likely reflecting the commitment of a more structured holiday.

The Tukey pairwise contrasts (Figure 11) show that not all segment differences are 
statistically distinguishable once multiple comparisons are accounted for. The most 
robust distinctions are between Corporate and leisure-oriented segments — a finding with 
direct implications for targeted marketing.

## Rate Ratio Forest Plot

```{r poisson-forest, fig.cap="Figure 10: Rate ratios from the quasi-Poisson GLM. Values above 1 indicate longer expected stays than the reference group. Error bars are 95% confidence intervals; blue points are significant at p < 0.05."}

pois_tidy <- tidy(model_pois, conf.int = TRUE) %>%
  mutate(
    rr      = exp(estimate),
    rr_low  = exp(conf.low),
    rr_high = exp(conf.high)
  ) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    term_clean  = term %>%
      str_replace("MarketSegment",    "Segment: ") %>%
      str_replace("CustomerType",     "Customer: ") %>%
      str_replace("arrival_season",   "Season: ") %>%
      str_replace("DepositType",      "Deposit: ") %>%
      str_replace("country_region",   "Region: ") %>%
      str_replace("Meal",             "Meal: ") %>%
      str_replace("avg_temp_c",       "Avg Temperature") %>%
      str_replace("precipitation_mm", "Precipitation") %>%
      str_replace("IsRepeatedGuest",  "Repeated Guest")
  ) %>%
  arrange(rr)

ggplot(pois_tidy, aes(x = rr, y = reorder(term_clean, rr), colour = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = rr_low, xmax = rr_high), height = 0.35, linewidth = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey30", linewidth = 0.8) +
  scale_colour_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey60"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"), name = "Significance"
  ) +
  scale_x_continuous(breaks = seq(0.4, 1.8, by = 0.2),
                     labels = function(x) paste0(x, "×")) +
  labs(
    title    = "Rate Ratios: Quasi-Poisson GLM for Total Nights Booked",
    subtitle = "Values > 1 = longer stays than reference; < 1 = shorter stays",
    x = "Rate Ratio [exp(β)]", y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

## Post-Hoc Contrasts: Which Segments Differ?

```{r poisson-glht, fig.cap="Figure 11: Tukey pairwise contrasts between market segments. Blue points indicate statistically significant differences after multiple comparison correction."}

glht_pois <- glht(model_pois, linfct = mcp(MarketSegment = "Tukey"))

glht_df <- as.data.frame(confint(glht_pois)$confint) %>%
  rownames_to_column("Contrast") %>%
  rename(estimate = Estimate, lower = lwr, upper = upr) %>%
  mutate(
    rr          = exp(estimate),
    rr_low      = exp(lower),
    rr_high     = exp(upper),
    significant = !(rr_low <= 1 & rr_high >= 1)
  )

ggplot(glht_df, aes(x = rr, y = reorder(Contrast, rr), colour = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = rr_low, xmax = rr_high), height = 0.4, linewidth = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "darkred", linewidth = 0.8) +
  scale_colour_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey60"),
    labels = c("TRUE" = "Significant (Tukey)", "FALSE" = "Not significant"), name = NULL
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "×")) +
  labs(
    title    = "Pairwise Market Segment Contrasts: Stay Length",
    subtitle = "Tukey-corrected; rate ratio > 1 means first segment stays longer",
    x = "Rate Ratio (exp[contrast])", y = NULL
  ) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top")
```

## Business Takeaway

```{r poisson-test-performance}
pred_pois <- predict(model_pois, newdata = test_pois, type = "response")
mae_pois  <- mean(abs(pred_pois - test_pois$total_stays))
rmse_pois <- sqrt(mean((pred_pois - test_pois$total_stays)^2))
cat(sprintf("Mean Absolute Error: %.2f nights\n", mae_pois))
cat(sprintf("RMSE:                %.2f nights\n", rmse_pois))
```


The quasi-Poisson model predicts stay length to within approximately `r round(mae_pois, 1)` nights on the held-out test set — a useful level of accuracy for segment-level planning. The business implication is straightforward: the hotel should prioritise channels and customer types associated with longer stays when capacity allows. Online TA and leisure segments in Summer represent the highest-value long-stay profiles. A targeted promotion, such as a discounted third or fourth night for Online TA bookings made more than 60 days in advance — could meaningfully shift the average stay length upward for this segment, generating incremental revenue with minimal additional operational cost.
---
# Model 3: Binomial GLM — Which Bookings Will Cancel?

*Section lead: Lorena Zovkic*

## Research Question

> **Which characteristics of a booking, known at the moment of reservation, most reliably predict whether the guest will cancel?**

Cancellations are the single largest source of revenue uncertainty for the hotel. A booking that cancels generates no revenue but consumes capacity that could have been sold to another guest. Identifying high-risk bookings at the point of reservation — before any room is allocated — allows the hotel to act: send retention communications, adjust overbooking levels, or flag reservations for manual review. This model builds that early-warning signal.

## Why This Model

`IsCanceled` is a binary outcome — a booking either cancels or it does not. The Binomial Generalised Linear Model (logistic regression) is the natural choice for binary response variables. It models the log-odds of cancellation as a linear function of predictors, guaranteeing predictions that remain within the [0, 1] probability range and producing coefficients interpretable as odds ratios. Crucially, only information available at the time of booking is used as predictors — the model is designed to score a reservation the moment it is made, not after the fact.

## Data Preparation

```{r binomial-prep}
hotel_bin <- hotel_data %>%
  filter(!as.character(MarketSegment) %in% c("Undefined", "Complementary"))
hotel_bin$MarketSegment <- droplevels(hotel_bin$MarketSegment)

set.seed(2026)
idx_bin   <- createDataPartition(hotel_bin$IsCanceled, p = 0.8, list = FALSE)
train_bin <- hotel_bin[idx_bin, ]
test_bin  <- hotel_bin[-idx_bin, ]

cat(sprintf("Training: %d | Test: %d | Cancel rate (train): %.1f%%\n",
            nrow(train_bin), nrow(test_bin),
            mean(train_bin$IsCanceled) * 100))
```

The binary outcome is well-balanced at 41.7% cancellations, meaning neither class dominates and standard threshold-based evaluation (accuracy, recall, precision) remains meaningful without resampling corrections. The same train/test split is used for the GAM in the following section, ensuring a fair comparison between the two models.

## Exploratory Plots

```{r binomial-eda, fig.cap="Figure 12: Cancellation rate by deposit type (left) and estimated cancellation probability as a function of lead time (right). The near-100% cancellation rate for Non Refund deposits is a data artefact discussed below. The LOESS smoother on lead time reveals a nonlinear pattern — risk rises steeply up to approximately 150 days before levelling off — which directly motivates the GAM in the next section.", fig.width=10, fig.height=4}

p_dep <- train_bin %>%
  group_by(DepositType) %>%
  summarise(rate = mean(IsCanceled), n = n(), .groups = "drop") %>%
  ggplot(aes(x = reorder(DepositType, rate), y = rate, fill = DepositType)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = paste0(scales::percent(rate, accuracy = 0.1),
                               "\n(n=", scales::comma(n), ")")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("No Deposit" = "steelblue",
                               "Non Refund" = "tomato",
                               "Refundable" = "darkorange")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1.25)) +
  labs(title = "Cancellation Rate by Deposit Type",
       subtitle = "Non Refund = 99.8% — likely a recording artefact",
       x = NULL, y = "Cancellation Rate") +
  theme_minimal(base_size = 10)

p_lead <- train_bin %>%
  ggplot(aes(x = LeadTime, y = IsCanceled)) +
  geom_smooth(method = "loess", span = 0.35,
              colour = "steelblue", fill = "lightblue", linewidth = 1.2) +
  geom_hline(yintercept = mean(train_bin$IsCanceled),
             linetype = "dashed", colour = "darkred", linewidth = 0.7) +
  annotate("text", x = 420, y = mean(train_bin$IsCanceled) + 0.04,
           label = paste0("Overall rate = ",
                          scales::percent(mean(train_bin$IsCanceled), accuracy = 0.1)),
           colour = "darkred", size = 3.2) +
  scale_x_continuous(limits = c(0, 500)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.9)) +
  labs(title = "Cancellation Probability by Lead Time",
       subtitle = "LOESS smoother — clearly nonlinear beyond ~150 days",
       x = "Lead Time (days before arrival)",
       y = "P(Cancellation)") +
  theme_minimal(base_size = 10)

p_dep + p_lead
```

## Model Fitting

```{r binomial-fit, cache=TRUE}
model_binomial <- glm(
  IsCanceled ~ LeadTime + DepositType + MarketSegment +
    PreviousCancellations + TotalOfSpecialRequests +
    IsRepeatedGuest + BookingChanges + DaysInWaitingList +
    CustomerType + is_via_agent + arrival_season,
  data   = train_bin,
  family = binomial
)

cat(sprintf("Null deviance:      %.1f on %d df\n",
            model_binomial$null.deviance, model_binomial$df.null))
cat(sprintf("Residual deviance:  %.1f on %d df\n",
            deviance(model_binomial), df.residual(model_binomial)))
cat(sprintf("McFadden pseudo-R²: %.3f\n",
            1 - deviance(model_binomial) / model_binomial$null.deviance))
```

## Formal Predictor Testing with drop1()

As in the Linear Model, we use `drop1()` with a Chi-square test to assess the contribution of each predictor to the overall model. For a Binomial GLM, this tests whether the deviance increases significantly when a predictor is removed — a more rigorous measure of importance than individual coefficient p-values, particularly for multi-level factors.

```{r binomial-drop1, fig.cap="Figure 13: Chi-square statistics from drop1() for each predictor in the Binomial GLM. The red dashed line marks the critical value at α = 0.05. All predictors cross the threshold, confirming that every variable in the model contributes significantly to predicting cancellation.", fig.height=5}

drop1_bin <- drop1(model_binomial, test = "Chisq")

drop1_bin_df <- as.data.frame(drop1_bin) %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "<none>") %>%
  mutate(Predictor = dplyr::recode(Predictor,
    "LeadTime"               = "Lead Time",
    "DepositType"            = "Deposit Type",
    "MarketSegment"          = "Market Segment",
    "PreviousCancellations"  = "Previous Cancellations",
    "TotalOfSpecialRequests" = "Special Requests",
    "IsRepeatedGuest"        = "Repeated Guest",
    "BookingChanges"         = "Booking Changes",
    "DaysInWaitingList"      = "Days on Waiting List",
    "CustomerType"           = "Customer Type",
    "is_via_agent"           = "Booked via Agent",
    "arrival_season"         = "Arrival Season"
  ))

crit_chi <- qchisq(0.95, df = 1)

ggplot(drop1_bin_df,
       aes(x = `LRT`, y = reorder(Predictor, `LRT`))) +
  geom_segment(aes(x = 0, xend = `LRT`,
                   yend = reorder(Predictor, `LRT`)),
               colour = "grey75", linewidth = 0.8) +
  geom_point(aes(colour = `Pr(>Chi)` < 0.05), size = 4) +
  scale_colour_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey50"),
    labels = c("TRUE" = "Significant (p < 0.05)", "FALSE" = "Not significant"),
    name   = NULL
  ) +
  geom_vline(xintercept = crit_chi, linetype = "dashed",
             colour = "darkred", linewidth = 0.8) +
  annotate("text", x = crit_chi + 2, y = 1,
           label = paste0("Critical χ² = ", round(crit_chi, 1)),
           colour = "darkred", size = 3.2, hjust = 0) +
  labs(
    title    = "Predictor Importance: drop1() Chi-square Statistics",
    subtitle = "Each bar shows how much deviance increases when that predictor is removed",
    x = "Likelihood Ratio Statistic (χ²)", y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

Every predictor in the model contributes significantly. The three strongest predictors — previous cancellations, deposit type, and lead time — are also the most actionable: they are all known at booking time and directly usable in a scoring system.

## Coefficient Interpretation

```{r binomial-forest, fig.cap="Figure 14: Odds ratios from the Binomial GLM with 95% confidence intervals. Values above 1 increase cancellation odds; values below 1 reduce them. The Non Refund deposit is excluded from this plot due to near-perfect separation (OR ≈ ∞) — discussed separately in the text. All predictors are shown on a log scale so that multiplicative effects appear symmetric.", fig.height=7, fig.width=8}

bin_tidy <- tidy(model_binomial, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)",
         !grepl("Non.Refund", term)) %>%
  mutate(
    significant = p.value < 0.05,
    term_clean  = term %>%
      str_replace("MarketSegment",         "Segment: ") %>%
      str_replace("CustomerType",          "Customer: ") %>%
      str_replace("DepositType",           "Deposit: ") %>%
      str_replace("arrival_season",        "Season: ") %>%
      str_replace("LeadTime",              "Lead Time") %>%
      str_replace("PreviousCancellations", "Previous Cancellations") %>%
      str_replace("TotalOfSpecialRequests","Special Requests") %>%
      str_replace("IsRepeatedGuest",       "Repeated Guest") %>%
      str_replace("BookingChanges",        "Booking Changes") %>%
      str_replace("DaysInWaitingList",     "Days on Waiting List") %>%
      str_replace("is_via_agent",          "Booked via Agent")
  )

ggplot(bin_tidy, aes(x = estimate,
                     y = reorder(term_clean, estimate),
                     colour = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.35, linewidth = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = "grey30", linewidth = 0.8) +
  scale_x_log10(labels = function(x) paste0(x, "×")) +
  scale_colour_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "grey60"),
    labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"), name = NULL
  ) +
  labs(
    title    = "Binomial GLM — Odds Ratios for Cancellation",
    subtitle = "Non Refund deposit excluded (OR ≈ ∞ — artefact, see text)",
    x = "Odds Ratio [exp(β)] — log scale", y = NULL
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Four predictors stand out for practical interpretation:

- **Lead time**: Each additional 10 days of lead time multiplies cancellation odds by approximately `r round(exp(coef(model_binomial)["LeadTime"] * 10), 2)`×. A booking made 200 days in advance carries roughly `r round(exp(coef(model_binomial)["LeadTime"] * 200) / exp(coef(model_binomial)["LeadTime"] * 7), 1)` times higher cancellation odds than one made one week out — the single most actionable signal available at booking time.

- **Special requests**: Each additional special request multiplies cancellation odds by `r round(exp(coef(model_binomial)["TotalOfSpecialRequests"]), 2)`×, reducing them by `r round((1 - exp(coef(model_binomial)["TotalOfSpecialRequests"])) * 100, 0)`%. Guests who invest in personalising their stay demonstrate genuine commitment — this is the strongest protective factor in the model.

- **Previous cancellations**: Each prior cancellation multiplies current odds by `r round(exp(coef(model_binomial)["PreviousCancellations"]), 2)`×. Booking history is a powerful and underused risk signal that requires no additional data collection.

- **Non Refund deposit** (excluded from plot): The estimated odds ratio is effectively infinite — 99.8% of such bookings are recorded as cancelled. This almost certainly reflects a recording convention where no-shows are retroactively reclassified rather than genuine guest behaviour. In a real deployment this variable would be excluded from the scoring model entirely.

## Marginal Effect: How Does Lead Time Translate Into Cancellation Risk?

```{r binomial-marginal, fig.cap="Figure 15: Predicted cancellation probability as a function of lead time for a typical reference booking (Online TA, Transient, No Deposit, Summer, no prior cancellations, one special request, booked via agent). The ribbon shows the 95% confidence interval. A booking made 200 days in advance has a cancellation probability roughly 25 percentage points higher than one made one week out — a practically significant difference for overbooking decisions.", fig.width=8, fig.height=4}

lead_seq_bin <- seq(0, 400, by = 2)

newdata_bin <- data.frame(
  LeadTime               = lead_seq_bin,
  DepositType            = factor("No Deposit",
                                  levels = levels(train_bin$DepositType)),
  MarketSegment          = factor("Online TA",
                                  levels = levels(train_bin$MarketSegment)),
  PreviousCancellations  = 0,
  TotalOfSpecialRequests = 1,
  IsRepeatedGuest        = 0L,
  BookingChanges         = 0,
  DaysInWaitingList      = 0,
  CustomerType           = factor("Transient",
                                  levels = levels(train_bin$CustomerType)),
  is_via_agent           = 1L,
  arrival_season         = factor("Summer",
                                  levels = levels(train_bin$arrival_season))
)

preds_bin <- predict(model_binomial,
                     newdata = newdata_bin,
                     type    = "link",
                     se.fit  = TRUE)

newdata_bin$prob     <- plogis(preds_bin$fit)
newdata_bin$prob_lwr <- plogis(preds_bin$fit - 1.96 * preds_bin$se.fit)
newdata_bin$prob_upr <- plogis(preds_bin$fit + 1.96 * preds_bin$se.fit)

ggplot(newdata_bin, aes(x = LeadTime)) +
  geom_ribbon(aes(ymin = prob_lwr, ymax = prob_upr),
              fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = prob), colour = "steelblue", linewidth = 1.3) +
  geom_hline(yintercept = mean(train_bin$IsCanceled),
             linetype = "dashed", colour = "darkred", linewidth = 0.7) +
  annotate("text", x = 320,
           y = mean(train_bin$IsCanceled) + 0.03,
           label = paste0("Overall cancel rate = ",
                          scales::percent(mean(train_bin$IsCanceled),
                                          accuracy = 0.1)),
           colour = "darkred", size = 3.2) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 400)) +
  labs(
    title    = "Marginal Effect of Lead Time on Cancellation Probability",
    subtitle = "Reference: Online TA · Transient · No Deposit · Summer · 1 special request",
    x        = "Lead Time (days before arrival)",
    y        = "Predicted Cancellation Probability"
  ) +
  theme_minimal(base_size = 11)
```

## Model Evaluation

```{r binomial-roc, fig.cap="Figure 16: ROC curve for the Binomial GLM on the held-out test set. The AUC measures the model's ability to correctly rank a random cancellation above a random non-cancellation, independent of the decision threshold chosen.", fig.width=6, fig.height=5}

pred_probs_bin <- predict(model_binomial, newdata = test_bin, type = "response")
roc_bin        <- roc(test_bin$IsCanceled, pred_probs_bin, quiet = TRUE)

plot(roc_bin,
     col  = "steelblue", lwd = 2,
     main = paste0("Binomial GLM — ROC Curve  (AUC = ",
                   round(auc(roc_bin), 3), ")"))
abline(a = 0, b = 1, lty = 2, col = "grey60")
```

```{r binomial-confmat}
pred_class_bin <- factor(ifelse(pred_probs_bin > 0.5, 1, 0))
cm_bin <- caret::confusionMatrix(pred_class_bin,
                                  factor(test_bin$IsCanceled),
                                  positive = "1")
cm_bin
```

## Business Takeaway

The Binomial GLM achieves an AUC of `r round(auc(roc_bin), 3)` on the held-out test set, using only information available at the moment of booking. For a revenue manager, three findings translate directly into operational decisions. First, any booking combining a lead time above 100 days, zero special requests, and at least one prior cancellation falls into the highest-risk tier and should trigger a proactive retention action — a personalised pre-arrival email or a flexible rate offer. Second, the marginal effect plot (Figure 15) provides a concrete basis for tiered overbooking: bookings made more than 150 days in advance justify a higher overbooking allowance than last-minute reservations. Third, the protective effect of special requests suggests a simple intervention: prompting guests to submit a room or meal preference within 48 hours of booking could meaningfully reduce cancellation rates at near-zero operational cost.

---

# Model 4: Generalised Additive Model — Are Cancellation Effects Truly Linear?

*Section lead: Lorena Zovkic*

## Research Question

> **Does cancellation risk accumulate linearly with lead time and room price — or does the relationship curve in ways a standard logistic regression cannot capture?**

The LOESS smoother in Figure EDA-5 and the exploratory plot in Figure 12 both show that cancellation risk rises steeply with lead time before plateauing around 150 days. A Binomial GLM must treat this as a straight line on the log-odds scale, which systematically misrepresents both the low-risk zone (short lead times) and the saturation zone (very long lead times). The GAM resolves this by letting the data determine the shape of each continuous relationship, rather than imposing linearity in advance.

## Why This Model

The GAM extends the Binomial GLM by replacing linear terms for continuous predictors with smooth spline functions. The complexity of each smooth is not fixed — it is selected automatically by penalised regression (REML), so the model uses only as much flexibility as the data supports. The key diagnostic is the **effective degrees of freedom (edf)**: an edf near 1 means the smooth collapsed to a straight line (the GLM assumption was justified); an edf substantially above 1 means the nonlinearity is real and the GLM missed it.

Because the GAM uses the same response variable, the same link function, and the same train/test split as the Binomial GLM, the two models are directly comparable.

## Model Fitting

```{r gam-fit, cache=TRUE}
library(mgcv)

model_gam <- gam(
  IsCanceled ~ s(LeadTime) + s(ADR) + s(DaysInWaitingList) +
    DepositType + MarketSegment + PreviousCancellations +
    TotalOfSpecialRequests + IsRepeatedGuest +
    CustomerType + arrival_season,
  data   = train_bin,
  family = binomial,
  method = "REML"
)

gam_summ <- summary(model_gam)
cat(sprintf("Deviance explained: %.1f%%\n", gam_summ$dev.expl * 100))
cat(sprintf("AIC:                %.1f\n",   AIC(model_gam)))
cat("\nSmooth term summary:\n")
print(round(gam_summ$s.table, 3))
```

## Smooth Term Plots

```{r gam-smooths, fig.cap="Figure 17: Estimated smooth terms from the GAM. Each panel shows how one continuous predictor affects the log-odds of cancellation after controlling for all other variables. The shaded band is a 95% confidence interval. The edf reported in each panel title quantifies how nonlinear the relationship is — edf = 1 would be a straight line.", fig.width=11, fig.height=4}

par(mfrow = c(1, 3), mar = c(4, 4, 4, 1))
plot(model_gam,
     shade      = TRUE,
     shade.col  = "lightblue",
     seWithMean = TRUE,
     scale      = 0,
     residuals  = FALSE,
     col        = "steelblue",
     lwd        = 2)
par(mfrow = c(1, 1))
```

Three patterns emerge from the smooth terms:

- **s(LeadTime)** has an edf of `r round(gam_summ$s.table["s(LeadTime)", "edf"], 1)`, confirming substantial nonlinearity. Cancellation risk rises steeply for bookings made up to approximately 100–150 days in advance, then plateaus. This plateau likely reflects highly committed long-planners — conference attendees, anniversary trips — who book early but rarely cancel. The Binomial GLM assigns a constantly increasing risk across this range and misrepresents both extremes.

- **s(ADR)** has an edf of `r round(gam_summ$s.table["s(ADR)", "edf"], 1)`. The relationship between room price and cancellation risk is not monotone — mid-range prices show the highest cancellation risk, while very cheap and very expensive rooms show lower risk. This pattern is invisible to the GLM.

- **s(DaysInWaitingList)** has an edf of `r round(gam_summ$s.table["s(DaysInWaitingList)", "edf"], 1)`, close to 1. For this predictor the GLM's linearity assumption holds — the small subset of bookings with waiting-list time behaves approximately linearly, and the extra flexibility of the GAM adds nothing here.

## GLM vs GAM: Does Flexibility Pay Off?

```{r gam-aic-compare}
aic_comparison <- AIC(model_binomial, model_gam)
print(round(aic_comparison, 1))
cat(sprintf("\nAIC improvement from GLM to GAM: %.1f points\n",
            AIC(model_binomial) - AIC(model_gam)))
```

```{r gam-roc, fig.cap="Figure 18: ROC curves for the Binomial GLM (blue) and GAM (red) evaluated on the same held-out test set. The vertical distance between the curves at any point represents the GAM's gain in sensitivity for a given false-positive rate.", fig.width=6, fig.height=5}

gam_probs <- predict(model_gam, newdata = test_bin, type = "response")
roc_gam   <- roc(test_bin$IsCanceled, gam_probs, quiet = TRUE)

plot(roc_bin,
     col  = "steelblue", lwd = 2,
     main = "ROC Curves: Binomial GLM vs GAM")
lines(roc_gam, col = "tomato", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = c(paste0("Binomial GLM (AUC = ", round(auc(roc_bin), 3), ")"),
                  paste0("GAM         (AUC = ", round(auc(roc_gam),  3), ")")),
       col = c("steelblue", "tomato"), lwd = 2, bty = "n")
```

```{r gam-confmat}
pred_class_gam <- factor(ifelse(gam_probs > 0.5, 1, 0))
cm_gam <- caret::confusionMatrix(pred_class_gam,
                                  factor(test_bin$IsCanceled),
                                  positive = "1")
cm_gam
```

## Marginal Smooth: Lead Time in Probability Units

```{r gam-marginal, fig.cap="Figure 19: Predicted cancellation probability from the GAM as a function of lead time (red), overlaid with the GLM prediction (blue) for the same reference booking. The GAM captures the plateau beyond 150 days that the GLM misses, and the steeper early rise below 50 days. These differences matter operationally — the GLM under-scores short-lead-time bookings and over-scores very long ones.", fig.width=8, fig.height=4}

newdata_gam <- data.frame(
  LeadTime               = lead_seq_bin,
  ADR                    = median(train_bin$ADR[train_bin$ADR > 0],
                                  na.rm = TRUE),
  DaysInWaitingList      = 0,
  DepositType            = factor("No Deposit",
                                  levels = levels(train_bin$DepositType)),
  MarketSegment          = factor("Online TA",
                                  levels = levels(train_bin$MarketSegment)),
  PreviousCancellations  = 0,
  TotalOfSpecialRequests = 1,
  IsRepeatedGuest        = 0L,
  CustomerType           = factor("Transient",
                                  levels = levels(train_bin$CustomerType)),
  arrival_season         = factor("Summer",
                                  levels = levels(train_bin$arrival_season))
)

gam_link   <- predict(model_gam, newdata = newdata_gam,
                      type = "link", se.fit = TRUE)
gam_margin <- data.frame(
  LeadTime = lead_seq_bin,
  prob     = plogis(gam_link$fit),
  prob_lwr = plogis(gam_link$fit - 1.96 * gam_link$se.fit),
  prob_upr = plogis(gam_link$fit + 1.96 * gam_link$se.fit)
)

glm_margin <- data.frame(
  LeadTime = lead_seq_bin,
  prob     = newdata_bin$prob
)

ggplot() +
  geom_ribbon(data = gam_margin,
              aes(x = LeadTime, ymin = prob_lwr, ymax = prob_upr),
              fill = "tomato", alpha = 0.15) +
  geom_line(data = gam_margin,
            aes(x = LeadTime, y = prob, colour = "GAM"),
            linewidth = 1.3) +
  geom_line(data = glm_margin,
            aes(x = LeadTime, y = prob, colour = "Binomial GLM"),
            linewidth = 1.1, linetype = "dashed") +
  scale_colour_manual(
    values = c("GAM" = "tomato", "Binomial GLM" = "steelblue"),
    name   = NULL
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 400)) +
  labs(
    title    = "GAM vs GLM: Predicted Cancellation Probability by Lead Time",
    subtitle = "Reference: Online TA · Transient · No Deposit · Summer · 1 special request · median ADR",
    x        = "Lead Time (days before arrival)",
    y        = "Predicted Cancellation Probability"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

## Business Takeaway

The GAM achieves an AUC of `r round(auc(roc_gam), 3)` versus `r round(auc(roc_bin), 3)` for the Binomial GLM — an improvement of `r round(auc(roc_gam) - auc(roc_bin), 3)` AUC points on held-out data. The AIC drops by `r round(AIC(model_binomial) - AIC(model_gam), 0)` points, confirming that the flexibility is genuinely justified and not overfitting. The practical implication is most visible in Figure 19: the GLM assigns a steadily increasing cancellation score across all lead times, while the GAM correctly identifies that risk stabilises beyond 150 days. A hotel using the GLM as a scoring engine would flag a 300-day booking at the same risk level as a 200-day booking — the GAM correctly treats them as equivalent, freeing up overbooking capacity that the GLM would unnecessarily reserve. For a risk-scoring system deployed in production, the GAM's additional complexity is well worth the modest implementation overhead.
---
# Model 5: Support Vector Machine

*Section lead: [Teammate name]*

<!-- Teammate section goes here -->

---

# Model 6: Neural Network

*Section lead: [Teammate name]*

<!-- Teammate section goes here -->

---

# Model Comparison

<!-- Shared section — to be completed together -->

---

# Conclusion

<!-- Shared section — to be completed together -->

---

# Challenges and Limitations

One notable data pipeline dependency warrants discussion. The weather enrichment relies on a live API call to the Open-Meteo historical weather archive at knit time. If the API is unavailable during rendering, the join produces missing values across all weather-derived predictors (`avg_temp_c`, `precipitation_mm`, `sunshine_hours`), causing downstream model chunks to fail. A production-grade pipeline would cache the API response locally after the first successful fetch, ensuring reproducibility independent of network access. This limitation is relevant to any team seeking to reproduce or extend this analysis.

---

# AI Usage Statement

<!-- Required by assignment — describe how AI tools were used in this project -->

---

# References

Antonio, N., de Almeida, A., & Nunes, L. (2019). Hotel booking demand datasets. *Data in Brief*, 22, 41–49.

Open-Meteo Historical Weather API. Retrieved from https://open-meteo.com
