---
title: "Predicting Hotel Booking Outcomes Using Machine Learning"
author: "Petronela Tabalae, Divya Shori, Lorena Zovkic"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
    df_print: paged
---

```{r libraries-setup, include = FALSE, cache = FALSE}
# NOTE: this chunk is intentionally NOT cached — cached chunks would not
# reload packages on subsequent renders, leading to "could not find function" errors.
library(tidyverse)
library(readr)
library(httr)
library(jsonlite)
library(dplyr)
library(lubridate)
library(car)
library(multcomp)
library(broom)
library(kableExtra)
library(scales)
library(patchwork)
library(caret)
library(pROC)

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  cache = TRUE
)
```

```{r load-hotel-data, include = FALSE}
hotel_data <- read_csv("H2.csv", na = c("NULL", "NA", ""), trim_ws = TRUE)
```

```{r load-weather-data, include = FALSE}
weather_cache_path <- "weather_raw.rds"

if (file.exists(weather_cache_path)) {
  weather_raw <- readRDS(weather_cache_path)
} else {
  url <- paste0(
    "https://archive-api.open-meteo.com/v1/archive",
    "?latitude=38.72&longitude=-9.13",
    "&start_date=2015-01-01&end_date=2017-12-31",
    "&daily=temperature_2m_mean,precipitation_sum,sunshine_duration",
    "&timezone=Europe/Lisbon"
  )
  response    <- GET(url)
  weather_raw <- fromJSON(content(response, "text", encoding = "UTF-8"))
  saveRDS(weather_raw, weather_cache_path)
}

weather_daily <- as_tibble(weather_raw$daily) |>
  mutate(time = as.Date(time))

weather_monthly <- weather_daily |>
  mutate(
    year  = year(time),
    month = month.name[month(time)]
  ) |>
  group_by(year, month) |>
  summarise(
    avg_temp_c       = mean(temperature_2m_mean, na.rm = TRUE),
    precipitation_mm = sum(precipitation_sum,    na.rm = TRUE),
    sunshine_hours   = sum(sunshine_duration,    na.rm = TRUE) / 3600,
    .groups = "drop"
  )

hotel_data <- hotel_data %>%
  left_join(weather_monthly,
            by = c("ArrivalDateYear" = "year",
                   "ArrivalDateMonth" = "month"))
```

```{r preprocessing, message = FALSE, warning = FALSE}
hotel_data <- hotel_data %>%
  mutate(
    across(where(is.character), trimws),
    Agent   = ifelse(Agent   %in% c("NULL", ""), NA, Agent),
    Company = ifelse(Company %in% c("NULL", ""), NA, Company),
    total_stays = StaysInWeekendNights + StaysInWeekNights,
    arrival_season = case_when(
      ArrivalDateMonth %in% c("March", "April", "May")            ~ "Spring",
      ArrivalDateMonth %in% c("June", "July", "August")           ~ "Summer",
      ArrivalDateMonth %in% c("September", "October", "November") ~ "Autumn",
      TRUE                                                          ~ "Winter"
    ),
    is_via_agent = ifelse(is.na(Agent), 0L, 1L),
    country_region = case_when(
      Country %in% c("PRT","ESP","FRA","ITA","DEU","BEL",
                     "NLD","CHE","AUT","LUX","AND")               ~ "Western Europe",
      Country %in% c("GBR","IRL","DNK","SWE","NOR","FIN",
                     "ISL","POL","CZE","SVK","HUN")               ~ "N/E Europe",
      Country %in% c("USA","CAN","BRA","ARG","MEX","COL")         ~ "Americas",
      TRUE                                                          ~ "Other"
    ),
    Children = ifelse(is.na(Children), 0, Children)
  )

hotel_data <- hotel_data[, setdiff(names(hotel_data),
                          c("ReservationStatus", "ReservationStatusDate", "Company"))]

hotel_data <- hotel_data %>%
  mutate(
    MarketSegment    = relevel(factor(MarketSegment),  ref = "Online TA"),
    CustomerType     = relevel(factor(CustomerType),   ref = "Transient"),
    DepositType      = relevel(factor(DepositType),    ref = "No Deposit"),
    Meal             = relevel(factor(Meal),           ref = "BB"),
    arrival_season   = relevel(factor(arrival_season), ref = "Summer"),
    country_region   = relevel(factor(country_region), ref = "Western Europe"),
    ReservedRoomType = factor(ReservedRoomType),
    IsCanceled       = as.integer(IsCanceled),
    IsRepeatedGuest  = as.integer(IsRepeatedGuest)
  )

mean_lt   <- round(mean(hotel_data$LeadTime), 1)
mean_temp <- round(mean(hotel_data$avg_temp_c, na.rm = TRUE), 1)

cat("Dataset ready:", nrow(hotel_data), "bookings,", ncol(hotel_data), "variables\n")
```

---

# Introduction

A hotel room unsold tonight is revenue gone forever. Unlike a retailer who can sell yesterday's inventory today, a hotel cannot recover a night that passed empty. This makes forecasting — of price, of stay length, of cancellations — not a nice-to-have but an operational necessity.

This report analyses **79,330 booking records** from a single city hotel in Lisbon, Portugal, covering reservations made between July 2015 and August 2017. The data were published by Antonio, Almeida, and Nunes (2019) and capture the full lifecycle of each booking: when it was made, who made it, what was requested, what was paid, and whether the guest ultimately arrived. Monthly weather observations for Lisbon — temperature, precipitation, and sunshine hours — from the Open-Meteo historical archive are joined as complementary predictors.

Three questions drive the analysis, each with direct operational stakes:

> **RQ1 — Price:** Which booking and seasonal characteristics determine the average daily rate a guest pays?

> **RQ2 — Stay length:** Which guest segments book the longest stays, and by how much?

> **RQ3 — Cancellation:** What information available at booking time most reliably predicts whether a reservation will be cancelled?

## Research Questions and Modelling Strategy

Six models address these questions. RQ1 and RQ2 each warrant a single well-specified parametric model: the response variables are continuous and count-valued respectively, and the relationships involved are well-described by additive structure on the appropriate transformed scale. RQ3 is approached through four models of increasing flexibility, because cancellation risk is the highest-stakes prediction problem and the one most likely to contain nonlinear and interactive structure that simpler models miss.

```{r modelling-ladder, echo = FALSE}
ladder_df <- data.frame(
  Model = c(
    "Model 1 — Linear Model",
    "Model 2 — Poisson GLM",
    "Model 3 — Binomial GLM",
    "Model 4 — GAM",
    "Model 5 — SVM (linear & RBF kernel)",
    "Model 6 — Neural Network"
  ),
  RQ = c("RQ1 — Price", "RQ2 — Stay length",
         "RQ3 — Cancellation", "RQ3 — Cancellation",
         "RQ3 — Cancellation", "RQ3 — Cancellation"),
  Rationale = c(
    "log(ADR) is approximately Gaussian; coefficients are multiplicative percentage effects on price — the language revenue managers use.",
    "total_stays is a non-negative integer count; Poisson GLM guarantees non-negative predictions and produces rate ratios. Overdispersion is tested and corrected if present.",
    "Linear log-odds baseline. Coefficients are odds ratios — interpretable and directly actionable. Establishes which predictors matter and in which direction.",
    "Replaces linear terms for three continuous predictors with penalised splines. Each smooth's effective degrees of freedom (EDF) quantifies how nonlinear the relationship truly is.",
    "The RBF kernel can represent arbitrary nonlinear decision boundaries and feature interactions. The linear kernel serves as an explicit control: any AUC gap between the two kernels isolates the contribution of nonlinearity from differences in fitting principle.",
    "Hidden units learn their own basis functions end-to-end. The most flexible model in the report; permutation importance reveals which predictors the network uses differently from the GLM."
  ),
  stringsAsFactors = FALSE
)

kable(ladder_df,
      caption = "Table 0: Six models across three research questions. Models 3–6 form a deliberate ladder: each step removes one specific assumption made by the step above, so the AUC gain at each rung has a clean methodological interpretation.",
      col.names = c("Model", "Research Question", "Rationale & what it adds"),
      align = c("l", "l", "l")) %>%
  kable_styling(full_width = TRUE, font_size = 11,
                bootstrap_options = c("striped", "hover")) %>%
  column_spec(1, width = "22%", bold = TRUE) %>%
  column_spec(2, width = "16%") %>%
  column_spec(3, width = "62%")
```

All models are trained on held-out splits and evaluated on unseen test data. For the four cancellation models, the split is identical — any difference in test-set performance reflects the model class, not the data each model had access to.

---

# Exploratory Data Analysis

Good analysis begins with honest description. This section examines what the dataset contains, what patterns emerge without any modelling assumptions, and what data quality issues require attention. Every preprocessing and modelling decision in the sections that follow traces back to a finding here.

## The Data at a Glance

```{r eda-snapshot, echo = FALSE}
snapshot <- data.frame(
  Area = c("Who booked", "How it was booked", "What was booked",
           "What happened", "When they arrived", "Weather context"),
  Key_Variables = c(
    "CustomerType, Country, Adults, Children, IsRepeatedGuest",
    "MarketSegment, DistributionChannel, Agent, LeadTime, DepositType",
    "ReservedRoomType, Meal, TotalOfSpecialRequests, RequiredCarParkingSpaces",
    "IsCanceled, ADR, StaysInWeekendNights, StaysInWeekNights",
    "ArrivalDateYear, ArrivalDateMonth, arrival_season",
    "avg_temp_c, precipitation_mm, sunshine_hours"
  ),
  Notes = c(
    "74.9% Transient; 39% from Portugal",
    "48.9% Online TA; median lead time 74 days",
    "BB meal plan = 78.5% of bookings",
    "41.7% cancelled; median ADR €100",
    "2015–2017; August is busiest month",
    "Joined from Open-Meteo API on arrival month"
  )
)

kable(snapshot,
      caption = "Table 1: Dataset structure — 79,330 bookings, 31 variables across six areas.",
      col.names = c("Area", "Key Variables", "Notable Facts")) %>%
  kable_styling(full_width = TRUE, font_size = 11) %>%
  row_spec(0, bold = TRUE)
```

Three variables are excluded before any model is fitted. `ReservationStatus` and `ReservationStatusDate` directly encode the cancellation outcome — including them would constitute data leakage. `Company` is missing for 94.7% of bookings and carries no usable signal.

## Seasonal Demand Patterns

```{r eda-temporal, fig.cap = "Figure EDA-1: Monthly booking volume across 2015–2017.", fig.height = 4}
hotel_data %>%
  mutate(
    ArrivalDateMonth = factor(ArrivalDateMonth, levels = month.name),
    Year = factor(ArrivalDateYear)
  ) %>%
  count(ArrivalDateMonth, Year) %>%
  ggplot(aes(x = ArrivalDateMonth, y = n, colour = Year, group = Year)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("2015" = "#2166ac",
                                  "2016" = "#4dac26",
                                  "2017" = "#d01c8b")) +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(title = "Monthly Booking Volume by Year",
       subtitle = "Consistent seasonal pattern across all three years",
       x = NULL, y = "Number of Bookings", colour = "Year") +
  theme_minimal(base_size = 11)
```

August dominates demand in every year while January and December are the quietest months. The seasonal rhythm is stable across 2015–2017 — not an artefact of one anomalous year. This stability has two modelling implications: arrival season is a reliable and meaningful predictor for all six models, and any model that ignores seasonality will systematically underperform.

## Distributions of Core Variables

```{r eda-distributions, fig.cap = "Figure EDA-2: Distributions of the four core numeric variables.", fig.height = 5}
p1 <- ggplot(hotel_data %>% filter(ADR > 0, ADR < 500), aes(x = ADR)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = median(hotel_data$ADR[hotel_data$ADR > 0]),
             linetype = "dashed", colour = "darkred", linewidth = 0.8) +
  scale_x_continuous(labels = dollar_format(prefix = "€")) +
  labs(title = "Room Price (ADR)", x = "€/night", y = "Count") +
  theme_minimal(base_size = 10)

p2 <- ggplot(hotel_data, aes(x = LeadTime)) +
  geom_histogram(bins = 60, fill = "darkorange", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = median(hotel_data$LeadTime),
             linetype = "dashed", colour = "darkred", linewidth = 0.8) +
  labs(title = "Lead Time", x = "Days before arrival", y = "Count") +
  theme_minimal(base_size = 10)

p3 <- hotel_data %>%
  filter(total_stays > 0, total_stays <= 12) %>%
  count(total_stays) %>%
  ggplot(aes(x = total_stays, y = n)) +
  geom_col(fill = "seagreen", colour = "white", alpha = 0.85) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Total Nights Booked", x = "Nights", y = "Count") +
  theme_minimal(base_size = 10)

p4 <- hotel_data %>%
  count(TotalOfSpecialRequests) %>%
  ggplot(aes(x = TotalOfSpecialRequests, y = n)) +
  geom_col(fill = "mediumpurple", colour = "white", alpha = 0.85) +
  scale_x_continuous(breaks = 0:5) +
  labs(title = "Special Requests", x = "Number of requests", y = "Count") +
  theme_minimal(base_size = 10)

(p1 + p2) / (p3 + p4)
```

ADR and lead time are both strongly right-skewed with long upper tails — this has direct modelling consequences. ADR must be log-transformed before serving as a response variable in the Linear Model; lead time's skew means a small number of bookings made 300–500 days in advance would exert disproportionate leverage on any linear fit. Both are centred before modelling so intercepts represent realistic baseline profiles. Stay length and special requests are discrete counts concentrated at low values, consistent with a Poisson distribution, confirming that count-regression methods are appropriate.

## Market Segment as the Primary Organising Variable

```{r eda-segment-overview, fig.cap = "Figure EDA-3: Mean ADR, mean stay length, and cancellation rate by market segment.", fig.height = 4}
seg_summary <- hotel_data %>%
  filter(!as.character(MarketSegment) %in% c("Undefined")) %>%
  group_by(MarketSegment) %>%
  summarise(
    mean_adr    = mean(ADR[ADR > 0], na.rm = TRUE),
    mean_stays  = mean(total_stays[total_stays > 0], na.rm = TRUE),
    cancel_rate = mean(IsCanceled),
    n           = n(),
    .groups = "drop"
  )

p_adr_seg <- ggplot(seg_summary,
                    aes(x = reorder(as.character(MarketSegment), mean_adr),
                        y = mean_adr,
                        fill = as.character(MarketSegment))) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = paste0("€", round(mean_adr))), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 145)) +
  labs(title = "Mean ADR", x = NULL, y = "€/night") +
  theme_minimal(base_size = 10)

p_stays_seg <- ggplot(seg_summary,
                      aes(x = reorder(as.character(MarketSegment), mean_adr),
                          y = mean_stays,
                          fill = as.character(MarketSegment))) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = round(mean_stays, 1)), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(limits = c(0, 5)) +
  labs(title = "Mean Stay", x = NULL, y = "Nights") +
  theme_minimal(base_size = 10)

p_cancel_seg <- ggplot(seg_summary,
                       aes(x = reorder(as.character(MarketSegment), mean_adr),
                           y = cancel_rate,
                           fill = as.character(MarketSegment))) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = scales::percent(cancel_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.95)) +
  labs(title = "Cancellation Rate", x = NULL, y = NULL) +
  theme_minimal(base_size = 10)

p_adr_seg + p_stays_seg + p_cancel_seg
```

Market segment simultaneously separates bookings by price, stay length, and cancellation risk — the three dimensions of this entire report. Direct bookings command the highest mean ADR (€122), reflecting that guests who book directly accept fewer intermediary discounts and tend to be more purposeful travellers. Groups sit at the opposite end on cancellation rate (69%), consistent with group bookings being tentative and subject to headcount changes. Online TA dominates volume at 48.9% of all bookings, making it the single most consequential segment for any revenue decision. Corporate guests stay the shortest, confirming that business travel generates revenue through price rather than nights.

## Two Data Quality Issues That Require Action

```{r eda-quality, fig.cap = "Figure EDA-4: Cancellation rate by deposit type (left) and ADR distribution highlighting the zero-price cluster (right).", fig.height = 4}
p_dep <- hotel_data %>%
  group_by(DepositType) %>%
  summarise(rate = mean(IsCanceled), n = n(), .groups = "drop") %>%
  ggplot(aes(x = reorder(DepositType, rate), y = rate, fill = DepositType)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = paste0(scales::percent(rate, accuracy = 0.1),
                               "\n(n=", scales::comma(n), ")")),
            hjust = -0.05, size = 3) +
  coord_flip() +
  scale_fill_manual(values = c("No Deposit"  = "steelblue",
                               "Non Refund"  = "tomato",
                               "Refundable"  = "darkorange")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1.2)) +
  labs(title = "Cancellation Rate by Deposit Type",
       subtitle = "Non Refund = 99.8% — likely a recording artefact",
       x = NULL, y = "Cancellation Rate") +
  theme_minimal(base_size = 10)

p_zero_adr <- hotel_data %>%
  mutate(zero_adr = ifelse(ADR == 0, "Zero ADR\n(complimentary)", "ADR > 0")) %>%
  filter(ADR < 500) %>%
  ggplot(aes(x = ADR, fill = zero_adr)) +
  geom_histogram(bins = 70, colour = "white", alpha = 0.85) +
  scale_fill_manual(values = c("Zero ADR\n(complimentary)" = "tomato",
                               "ADR > 0" = "steelblue"), name = NULL) +
  scale_x_continuous(labels = dollar_format(prefix = "€")) +
  labs(title = "ADR Distribution",
       subtitle = "1,208 zero-ADR bookings excluded from pricing model",
       x = "ADR (€/night)", y = "Count") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")

p_dep + p_zero_adr
```

Two findings here require explicit handling before any model is fitted. The 99.8% cancellation rate for Non Refund bookings is not a genuine behavioural signal — no hotel segment cancels at near certainty. The most credible explanation is a back-office recording convention where no-shows and agent-initiated failures are logged as "cancelled" regardless of payment status. This variable is included in all cancellation models because it genuinely separates booking populations and improves prediction, but its coefficient cannot be interpreted causally and should not be treated as a business lever.

The 1,208 zero-ADR bookings represent complimentary rooms — gifted stays not priced by the market. Including them would corrupt any model of room price by pulling the mean downward. They are excluded from Model 1 but retained in Models 3–6 where ADR is a predictor rather than the outcome.

## The Nonlinear Lead Time — Cancellation Relationship

```{r eda-leadtime, fig.cap = "Figure EDA-5: Estimated cancellation probability as a function of lead time (LOESS smoother, n = 10,000 random subsample).", fig.height = 4}
set.seed(2026)
hotel_data %>%
  slice_sample(n = 10000) %>%
  ggplot(aes(x = LeadTime, y = IsCanceled)) +
  geom_smooth(method = "loess", span = 0.35,
              colour = "steelblue", fill = "lightblue", linewidth = 1.2) +
  geom_hline(yintercept = mean(hotel_data$IsCanceled),
             linetype = "dashed", colour = "darkred", linewidth = 0.7) +
  annotate("text", x = 420, y = mean(hotel_data$IsCanceled) + 0.03,
           label = paste0("Overall rate = ",
                          scales::percent(mean(hotel_data$IsCanceled), accuracy = 0.1)),
           colour = "darkred", size = 3.2) +
  scale_x_continuous(limits = c(0, 500)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.85)) +
  labs(title = "Cancellation Probability by Lead Time",
       subtitle = "LOESS smoother — risk rises steeply, then plateaus after ~150 days",
       x = "Lead Time (days before arrival)",
       y = "Estimated Cancellation Probability") +
  theme_minimal(base_size = 11)
```

This is the single most consequential plot in the EDA. Cancellation risk rises steeply for bookings made up to approximately 150 days in advance — then clearly plateaus. A guest booking 300 days out is not meaningfully riskier than one booking 200 days out. This plateau likely reflects committed long-horizon planners: conference attendees, anniversary travellers, families coordinating around school calendars who book early by necessity, not indecision. A model that imposes a straight line through this curve — as the Binomial GLM must — will overestimate risk for this group and potentially underestimate it at the steepest part of the early rise. This single observation is the primary technical motivation for the GAM in Model 4.

## What the EDA Tells Us Before Any Model Is Fitted

Five decisions follow directly from this section and are applied consistently across all six models:

1. **Log-transform ADR** for the Linear Model — the right skew violates normality and would distort coefficient estimates.
2. **Exclude zero-ADR records** from pricing models — complimentary rooms are not market prices.
3. **Exclude zero-night stays** from the Poisson model — 331 bookings with no recorded stay cannot be meaningfully modelled as a count outcome.
4. **Centre continuous predictors** (lead time, temperature) so that model intercepts represent realistic baseline profiles rather than impossible values of zero days and zero degrees.
5. **Flag Non Refund deposit** in all cancellation models — include for prediction accuracy, never interpret causally.

---

# Model 1: Linear Model — What Drives Room Price?

*Section lead: Divya Shori*

## Research Question

> **Which booking and seasonal characteristics determine the average daily rate a guest pays, and by how much?**

Room price is the primary revenue lever in a hotel. Even small systematic differences across segments or seasons compound across thousands of bookings into significant revenue outcomes. This model builds the evidential basis for differential pricing: not a heuristic but a quantified, controlled estimate — "Direct bookings sustain a 12% price premium over Online TA, holding season, room type, and lead time constant."

## Why a Linear Model?

ADR is a continuous, positive amount. Once log-transformed to address right skew, it behaves approximately as a Gaussian variable on the linear scale, making the Linear Model the natural and principled choice. The log transformation carries a further advantage: coefficients become multiplicative percentage effects on price, which is the language revenue managers use. A coefficient of 0.10 means a 10% price premium. No other model class produces this combination of interpretability and correctness for a log-normal response.

## Data Preparation and Motivation for log(ADR)

```{r lm-motivation-plot, fig.cap = "Figure 1: Raw ADR (left) is strongly right-skewed. After log-transformation (right) the distribution is approximately symmetric — confirming that log(ADR) is the correct response variable.", fig.height = 4}
hotel_lm <- hotel_data %>%
  filter(ADR > 0,
         !as.character(MarketSegment) %in% c("Complementary", "Undefined")) %>%
  mutate(
    log_adr      = log(ADR),
    LeadTime_c   = LeadTime   - mean(LeadTime,   na.rm = TRUE),
    avg_temp_c_c = avg_temp_c - mean(avg_temp_c, na.rm = TRUE)
  )
hotel_lm$MarketSegment <- droplevels(hotel_lm$MarketSegment)

p_raw <- ggplot(hotel_lm, aes(x = ADR)) +
  geom_histogram(bins = 80, fill = "tomato", colour = "white", alpha = 0.85) +
  scale_x_continuous(labels = dollar_format(prefix = "€"), limits = c(0, 600)) +
  labs(title = "Raw ADR", subtitle = "Right-skewed — violates LM normality assumption",
       x = "ADR (€/night)", y = "Count") +
  theme_minimal(base_size = 11)

p_log <- ggplot(hotel_lm, aes(x = log_adr)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = mean(hotel_lm$log_adr),
             linetype = "dashed", colour = "black", linewidth = 0.8) +
  annotate("text", x = mean(hotel_lm$log_adr) + 0.15, y = Inf, vjust = 2,
           label = paste0("Mean = €", round(exp(mean(hotel_lm$log_adr)), 1)), size = 3.5) +
  labs(title = "log(ADR)", subtitle = "Approximately symmetric — suitable for LM",
       x = "log(ADR)", y = "Count") +
  theme_minimal(base_size = 11)

p_raw + p_log
```

The transformation is not cosmetic — it is a prerequisite for a valid Linear Model. On the log scale, every coefficient is a multiplicative percentage effect on price, directly comparable to how revenue managers think about pricing premiums. Two continuous predictors are mean-centred before fitting: lead time (mean = `r mean_lt` days) and monthly temperature (mean = `r mean_temp`°C). Centring ensures the model intercept has a concrete interpretation — the expected room price for the most common booking profile at average lead time and average temperature — rather than an extrapolation to impossible values.

`Online TA` is set as the reference level for market segment (49% of bookings) and `Transient` for customer type (75% of bookings), ensuring the intercept represents the most common booking profile.

## Exploring Key Predictors

```{r lm-eda, fig.cap = "Figure 2: log(ADR) by market segment (left) and by lead time (right).", fig.height = 4}
p_seg <- ggplot(hotel_lm,
                aes(x = reorder(MarketSegment, log_adr, median),
                    y = log_adr, fill = MarketSegment)) +
  geom_violin(trim = TRUE, alpha = 0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0("€", round(exp(x))),
                     sec.axis = sec_axis(~ ., name = "log(ADR)")) +
  labs(title = "Room Price by Market Segment", x = NULL, y = "ADR (€) [log scale]") +
  theme_minimal(base_size = 10)

p_lead <- ggplot(hotel_lm %>% sample_n(8000), aes(x = LeadTime, y = log_adr)) +
  geom_point(alpha = 0.12, size = 0.7, colour = "steelblue") +
  geom_smooth(method = "loess", span = 0.4,
              colour = "darkred", fill = "salmon", linewidth = 1.1) +
  scale_x_continuous(limits = c(0, 400)) +
  labs(title = "Lead Time vs. Room Price",
       x = "Lead Time (days before arrival)", y = "log(ADR)") +
  theme_minimal(base_size = 10)

p_seg + p_lead
```

Market segment is a strong price discriminator — Direct bookings sit clearly above Groups and Corporate. The lead time scatter shows a modest but consistent negative association: bookings made further in advance tend to achieve lower prices, consistent with early-bird discount strategies. Crucially, the LOESS smoother is approximately linear here — unlike the cancellation relationship in Figure EDA-5 — which justifies the Linear Model's treatment of lead time as a linear predictor for this response variable.

## Model Fitting

```{r lm-split, cache = TRUE}
set.seed(2026)
idx_lm   <- createDataPartition(hotel_lm$log_adr, p = 0.8, list = FALSE)
train_lm <- hotel_lm[idx_lm, ]
test_lm  <- hotel_lm[-idx_lm, ]
cat("Training:", nrow(train_lm), "| Test:", nrow(test_lm), "\n")
```

```{r lm-fit, cache = TRUE, dependson = "lm-split"}
model_lm <- lm(
  log_adr ~ arrival_season + MarketSegment + CustomerType +
            ReservedRoomType + LeadTime_c + TotalOfSpecialRequests +
            IsRepeatedGuest + Meal + is_via_agent + avg_temp_c_c,
  data = train_lm
)

cat(sprintf("Adjusted R²  : %.3f\n", summary(model_lm)$adj.r.squared))
cat(sprintf("Residual SE  : %.3f log-ADR units\n", summary(model_lm)$sigma))
```

## Formal Predictor Testing with drop1()

Rather than relying on individual coefficient p-values — which are unreliable for multi-level factors where no single coefficient captures the full effect — we apply `drop1()` with an F-test to assess the contribution of each predictor to the overall model.

```{r lm-drop1, dependson = "lm-fit", fig.cap = "Figure 3: F-statistics from drop1() for each predictor. The dashed line marks the critical F-value at α = 0.05.", fig.height = 5}
drop1_res <- drop1(model_lm, test = "F")

drop1_df <- as.data.frame(drop1_res) %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "<none>") %>%
  mutate(Predictor = dplyr::recode(Predictor,
    "arrival_season"         = "Arrival Season",
    "MarketSegment"          = "Market Segment",
    "CustomerType"           = "Customer Type",
    "ReservedRoomType"       = "Reserved Room Type",
    "LeadTime_c"             = "Lead Time (centred)",
    "TotalOfSpecialRequests" = "Special Requests",
    "IsRepeatedGuest"        = "Repeated Guest",
    "Meal"                   = "Meal Plan",
    "is_via_agent"           = "Booked via Agent",
    "avg_temp_c_c"           = "Avg Temperature (centred)"
  ))

crit_f <- qf(0.95, df1 = 1, df2 = df.residual(model_lm))

ggplot(drop1_df, aes(x = `F value`, y = reorder(Predictor, `F value`))) +
  geom_segment(aes(x = 0, xend = `F value`, yend = reorder(Predictor, `F value`)),
               colour = "grey75", linewidth = 0.8) +
  geom_point(aes(colour = `Pr(>F)` < 0.05), size = 4) +
  scale_colour_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey50"),
                      labels = c("TRUE" = "Significant (p < 0.05)", "FALSE" = "Not significant"),
                      name = NULL) +
  geom_vline(xintercept = crit_f, linetype = "dashed",
             colour = "darkred", linewidth = 0.8) +
  annotate("text", x = crit_f + 5, y = 1,
           label = paste0("Critical F = ", round(crit_f, 1)),
           colour = "darkred", size = 3.2, hjust = 0) +
  labs(title = "Predictor Importance: drop1() F-Statistics",
       subtitle = "Each bar shows how much model fit deteriorates when that predictor is removed",
       x = "F-statistic", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

Every predictor in the model clears the significance threshold — the model contains no dead weight. The ranking is directly informative: room type and market segment generate the largest F-statistics, confirming that price is primarily a function of what was booked and through which channel. Arrival season and meal plan follow closely, reflecting peak-demand premiums and the added value of inclusive packages. Lead time and temperature sit lower but remain significant — their effects are real, just smaller in magnitude relative to the dominant structural drivers.

## The Pure Effect of Lead Time: Added-Variable Plot

A simple scatter of log(ADR) against lead time conflates the lead time effect with everything else that varies with it — segments that tend to book far in advance, seasons with different pricing, and so on. An added-variable plot removes this confounding by showing only what lead time explains *after* every other predictor has already been accounted for.

```{r lm-avplot, dependson = "lm-fit", fig.cap = "Figure 4: Added-variable plot for lead time. Both axes are residuals after removing all other predictors. The negative slope is the pure, unconfounded effect of lead time on room price.", fig.height = 4}
model_no_lt <- lm(
  log_adr ~ arrival_season + MarketSegment + CustomerType +
            ReservedRoomType + TotalOfSpecialRequests +
            IsRepeatedGuest + Meal + is_via_agent + avg_temp_c_c,
  data = train_lm
)
model_lt_on_x <- lm(
  LeadTime_c ~ arrival_season + MarketSegment + CustomerType +
               ReservedRoomType + TotalOfSpecialRequests +
               IsRepeatedGuest + Meal + is_via_agent + avg_temp_c_c,
  data = train_lm
)

av_df <- data.frame(
  resid_y  = residuals(model_no_lt),
  resid_lt = residuals(model_lt_on_x)
)

ggplot(av_df %>% sample_n(6000), aes(x = resid_lt, y = resid_y)) +
  geom_point(alpha = 0.12, size = 0.7, colour = "steelblue") +
  geom_smooth(method = "lm", colour = "darkred", fill = "salmon",
              linewidth = 1.2, se = TRUE) +
  labs(title = "Added-Variable Plot: Lead Time → Room Price",
       subtitle = "Pure effect of lead time after controlling for all other predictors",
       x = "Lead Time residuals (days — unexplained by other predictors)",
       y = "log(ADR) residuals (unexplained by other predictors)") +
  theme_minimal(base_size = 11)
```

The negative slope is unambiguous and is not an artefact of confounding with season or market segment. After removing the influence of every other predictor in the model, guests who book earlier still pay less. The practical implication is direct: early-booking discounts are not simply a consequence of different guest profiles — they reflect a genuine pricing dynamic where advance purchase is rewarded, and the hotel can justify structured early-bird pricing tiers on this evidence.

## Marginal Effect: What Does a 200-Day Booking Cost?

```{r lm-marginal, dependson = "lm-fit", fig.cap = "Figure 5: Predicted ADR as a function of lead time for a typical reference booking. Shaded band = 95% confidence interval.", fig.height = 4}
lead_seq    <- seq(0, 400, by = 2)
mean_lt_val <- mean(train_lm$LeadTime, na.rm = TRUE)
mode_room   <- names(sort(table(train_lm$ReservedRoomType), decreasing = TRUE))[1]

newdata_lt <- data.frame(
  LeadTime_c             = lead_seq - mean_lt_val,
  MarketSegment          = factor("Online TA",  levels = levels(train_lm$MarketSegment)),
  CustomerType           = factor("Transient",  levels = levels(train_lm$CustomerType)),
  arrival_season         = factor("Summer",     levels = levels(train_lm$arrival_season)),
  ReservedRoomType       = factor(mode_room,    levels = levels(train_lm$ReservedRoomType)),
  Meal                   = factor("BB",         levels = levels(train_lm$Meal)),
  TotalOfSpecialRequests = mean(train_lm$TotalOfSpecialRequests, na.rm = TRUE),
  IsRepeatedGuest        = 0L,
  is_via_agent           = 1L,
  avg_temp_c_c           = 0
)

preds_lt       <- predict(model_lm, newdata = newdata_lt, interval = "confidence")
newdata_lt$fit <- exp(preds_lt[, "fit"])
newdata_lt$lwr <- exp(preds_lt[, "lwr"])
newdata_lt$upr <- exp(preds_lt[, "upr"])

ggplot(newdata_lt, aes(x = lead_seq)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "steelblue", alpha = 0.2) +
  geom_line(aes(y = fit), colour = "steelblue", linewidth = 1.3) +
  geom_vline(xintercept = mean_lt_val, linetype = "dashed",
             colour = "grey50", linewidth = 0.7) +
  annotate("text", x = mean_lt_val + 8, y = max(newdata_lt$upr) * 0.98,
           label = paste0("Mean lead time\n(", round(mean_lt_val), " days)"),
           size = 3.2, colour = "grey40", hjust = 0) +
  scale_y_continuous(labels = dollar_format(prefix = "€")) +
  labs(title = "Marginal Effect of Lead Time on Room Price",
       subtitle = "Reference: Online TA · Transient · Summer · BB meal · via agent",
       x = "Lead Time (days before arrival)",
       y = "Predicted ADR (€/night)") +
  theme_minimal(base_size = 11)
```

For the reference booking profile, the model predicts a price decline of roughly €20–25 across the full range from last-minute to 400 days ahead. A guest booking tomorrow pays around €20 more per night than one who booked six months ago — a meaningful difference at scale but not a dramatic one. Last-minute rates can be set at a modest premium without triggering significant demand loss, while early-bird discounts of 10–15% are consistent with the data.

## Model Diagnostics

```{r lm-diagnostics, dependson = "lm-fit", fig.cap = "Figure 6: Residual diagnostics. Broadly homoscedastic spread and approximate normality confirm the log-transformation was appropriate.", fig.height = 4}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(model_lm, which = 1, main = "Residuals vs Fitted",
     col = adjustcolor("steelblue", 0.3), pch = 16, cex = 0.5)
plot(model_lm, which = 2, main = "Normal Q-Q",
     col = adjustcolor("steelblue", 0.5), pch = 16, cex = 0.5)
par(mfrow = c(1, 1))
```

The residuals vs. fitted plot shows no strong systematic pattern — variance is broadly stable across the fitted range. The Q-Q plot confirms approximate normality through the central range, with modest tail deviations typical for datasets of this size. These diagnostics confirm that the log-transformation successfully addressed the skewness identified in Figure 1.

## Out-of-Sample Performance

```{r lm-test, dependson = "lm-fit"}
pred_lm  <- predict(model_lm, newdata = test_lm)
rmse_lm  <- sqrt(mean((pred_lm - test_lm$log_adr)^2))
r2_test  <- cor(pred_lm, test_lm$log_adr)^2
rmse_eur <- exp(mean(test_lm$log_adr) + rmse_lm) - exp(mean(test_lm$log_adr))

cat(sprintf("Adjusted R² (train): %.3f\n", summary(model_lm)$adj.r.squared))
cat(sprintf("R² on test set:      %.3f\n", r2_test))
cat(sprintf("Approx. error in €:  ±€%.1f per night\n", abs(rmse_eur)))
```

```{r lm-pred-vs-actual, dependson = "lm-test", fig.cap = "Figure 7: Predicted vs. observed ADR on the held-out test set. Points cluster tightly around the 45° perfect-prediction line.", fig.height = 4}
data.frame(predicted = pred_lm, observed = test_lm$log_adr) %>%
  sample_n(3000) %>%
  ggplot(aes(x = exp(predicted), y = exp(observed))) +
  geom_point(alpha = 0.2, size = 0.8, colour = "steelblue") +
  geom_abline(slope = 1, intercept = 0,
              colour = "darkred", linetype = "dashed", linewidth = 0.9) +
  scale_x_continuous(labels = dollar_format(prefix = "€"), limits = c(0, 600)) +
  scale_y_continuous(labels = dollar_format(prefix = "€"), limits = c(0, 600)) +
  labs(title = "Predicted vs. Observed ADR — Test Set",
       subtitle = "Red dashed line = perfect prediction",
       x = "Predicted ADR (€/night)", y = "Observed ADR (€/night)") +
  theme_minimal(base_size = 11)
```

## Business Takeaway

The Linear Model explains `r round(summary(model_lm)$adj.r.squared * 100, 1)`% of room price variance on the training set, with an out-of-sample R² of `r round(r2_test, 3)` and a practical prediction error of approximately ±€`r round(abs(rmse_eur), 1)` per night — confirming genuine out-of-sample generalisation without overfitting. Three findings translate directly into commercial decisions. First, market segment is the dominant price driver: Direct bookings sustain the highest rates, which justifies investing in direct channel marketing to shift the booking mix toward higher-margin customers. Second, Summer arrivals attract a meaningful seasonal premium; off-peak rate strategies should account for this baseline. Third, the negative lead-time effect provides a data-grounded argument for structured early-bird pricing tiers that reward advance commitment while protecting last-minute rate integrity.

---

# Model 2: Poisson GLM — Who Stays Longest?

*Section lead: Divya Shori*

## Research Question

> **Which guest segments and booking characteristics predict longer stays, and how large are the differences across segments?**

A guest who stays four nights generates roughly twice the revenue of a two-night guest, with proportionally lower per-night operational cost. Yet most channel and marketing strategies are optimised for price or cancellation rate, not stay length. This model quantifies which segments stay longest — and by how much — providing the evidential basis for targeted long-stay promotions.

## Why a Poisson GLM?

`total_stays` is a non-negative integer count. A Linear Model is inappropriate — it can predict negative values and treats fractional nights as equally plausible. The Poisson GLM models the log of the expected count as a linear function of predictors, guarantees non-negative predictions, and produces **rate ratios** directly interpretable as "segment X stays Y% longer than the reference." Before accepting the Poisson assumption, we formally test for overdispersion — the common real-world violation where variance exceeds the mean, causing Poisson standard errors to be underestimated and inference to be anti-conservative.

## Data Preparation

```{r poisson-dist, fig.cap = "Figure 8: Observed stay length (blue) vs. the theoretical Poisson distribution at the observed mean (red).", fig.height = 4}
hotel_pois <- hotel_data %>%
  filter(total_stays > 0,
         !as.character(MarketSegment) %in% c("Undefined"))
hotel_pois$MarketSegment <- droplevels(hotel_pois$MarketSegment)

obs_lambda  <- mean(hotel_pois$total_stays)
max_display <- 14

pois_theoretical <- data.frame(
  nights   = 1:max_display,
  expected = dpois(1:max_display, lambda = obs_lambda) * nrow(hotel_pois)
)

hotel_pois %>%
  filter(total_stays <= max_display) %>%
  ggplot(aes(x = total_stays)) +
  geom_histogram(aes(y = after_stat(count)), binwidth = 1,
                 fill = "steelblue", colour = "white", alpha = 0.75) +
  geom_line(data = pois_theoretical, aes(x = nights, y = expected),
            colour = "darkred", linewidth = 1.3) +
  geom_point(data = pois_theoretical, aes(x = nights, y = expected),
             colour = "darkred", size = 2.5) +
  scale_x_continuous(breaks = 1:max_display) +
  labs(title = "Distribution of Total Nights Booked",
       subtitle = paste0("Observed (blue) vs. Poisson(λ = ", round(obs_lambda, 2), ") (red)"),
       x = "Total Nights Booked", y = "Number of Bookings") +
  theme_minimal(base_size = 11)
```

The theoretical Poisson curve fits reasonably at low counts but falls well short of the observed frequencies at 4–8 nights. The data's right tail is heavier than a Poisson distribution predicts — a visual indicator that variance exceeds the mean. We confirm this formally below.

```{r poisson-meanvar, fig.cap = "Figure 9: Mean vs. variance of total nights across market segment × season groups. Points above the Var = Mean line confirm systematic overdispersion.", fig.height = 4}
mv_data <- hotel_pois %>%
  group_by(MarketSegment, arrival_season) %>%
  summarise(
    group_mean = mean(total_stays),
    group_var  = var(total_stays),
    n          = n(),
    .groups    = "drop"
  ) %>%
  filter(n >= 30)

ggplot(mv_data, aes(x = group_mean, y = group_var,
                    colour = MarketSegment, size = n)) +
  geom_point(alpha = 0.85) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "darkred", linewidth = 0.9) +
  annotate("text", x = 1.3, y = 1.2,
           label = "Var = Mean\n(Poisson assumption)", colour = "darkred",
           size = 3.2, hjust = 0) +
  scale_size_continuous(range = c(2, 9), name = "Bookings in group") +
  scale_colour_brewer(palette = "Set2", name = "Market Segment") +
  labs(title = "Mean–Variance Relationship by Segment and Season",
       subtitle = "Points above the dashed line signal overdispersion (Var > Mean)",
       x = "Group Mean (nights)", y = "Group Variance") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right")
```

Nearly every segment–season combination sits above the Var = Mean line. Overdispersion is not confined to one unusual subgroup — it is a structural feature of this data. Standard Poisson standard errors would be too narrow, leading to over-rejection of null hypotheses. The quasi-Poisson correction below addresses this by scaling standard errors upward while leaving coefficients unchanged.

## Model Fitting and Overdispersion Correction

```{r poisson-fit, cache = TRUE}
set.seed(2026)
idx_pois   <- createDataPartition(hotel_pois$total_stays, p = 0.8, list = FALSE)
train_pois <- hotel_pois[idx_pois, ]
test_pois  <- hotel_pois[-idx_pois, ]

model_pois <- glm(
  total_stays ~ MarketSegment + CustomerType + arrival_season +
                DepositType + IsRepeatedGuest + Meal +
                avg_temp_c + precipitation_mm + country_region,
  data = train_pois, family = poisson
)

disp_ratio <- deviance(model_pois) / df.residual(model_pois)
cat(sprintf("Dispersion ratio (Deviance / df): %.2f\n", disp_ratio))
cat(sprintf("→ %s\n",
    ifelse(disp_ratio > 1.5,
           "Overdispersion confirmed — refitting with quasi-Poisson.",
           "No strong overdispersion — Poisson retained.")))
```

```{r quasipoisson-refit, cache = TRUE, dependson = "poisson-fit"}
if (disp_ratio > 1.5) {
  model_pois <- glm(
    total_stays ~ MarketSegment + CustomerType + arrival_season +
                  DepositType + IsRepeatedGuest + Meal +
                  avg_temp_c + precipitation_mm + country_region,
    data = train_pois, family = quasipoisson
  )
  cat("Quasi-Poisson fitted. Coefficients unchanged; SEs corrected upward.\n")
}

model_null <- glm(total_stays ~ 1, data = train_pois, family = quasipoisson)
anova(model_null, model_pois, test = "F")
```

The dispersion ratio of `r round(disp_ratio, 2)` confirms severe overdispersion. The quasi-Poisson model corrects this by scaling standard errors upward by √`r round(disp_ratio, 2)` ≈ `r round(sqrt(disp_ratio), 2)`. Critically, the coefficients (rate ratios) are identical to standard Poisson — only the uncertainty estimates change, and they change in the direction of honesty. The F-test against the null model confirms that the predictors collectively explain significant variation in stay length.

## Rate Ratio Forest Plot

```{r poisson-forest, dependson = "quasipoisson-refit", fig.cap = "Figure 10: Rate ratios from the quasi-Poisson GLM with 95% confidence intervals. Values above 1× indicate longer expected stays than the reference group.", fig.height = 7}
pois_tidy <- tidy(model_pois, conf.int = TRUE) %>%
  mutate(
    rr      = exp(estimate),
    rr_low  = exp(conf.low),
    rr_high = exp(conf.high)
  ) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    significant = p.value < 0.05,
    term_clean  = term %>%
      str_replace("MarketSegment",    "Segment: ") %>%
      str_replace("CustomerType",     "Customer: ") %>%
      str_replace("arrival_season",   "Season: ") %>%
      str_replace("DepositType",      "Deposit: ") %>%
      str_replace("country_region",   "Region: ") %>%
      str_replace("Meal",             "Meal: ") %>%
      str_replace("avg_temp_c",       "Avg Temperature") %>%
      str_replace("precipitation_mm", "Precipitation") %>%
      str_replace("IsRepeatedGuest",  "Repeated Guest")
  ) %>%
  arrange(rr)

ggplot(pois_tidy, aes(x = rr, y = reorder(term_clean, rr), colour = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = rr_low, xmax = rr_high), height = 0.35, linewidth = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey30", linewidth = 0.8) +
  scale_colour_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey60"),
                      labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
                      name = "Significance") +
  scale_x_continuous(breaks = seq(0.4, 1.8, by = 0.2),
                     labels = function(x) paste0(x, "×")) +
  labs(title = "Rate Ratios: Quasi-Poisson GLM for Total Nights Booked",
       subtitle = "Values > 1× = longer stays than reference; < 1× = shorter",
       x = "Rate Ratio [exp(β)]", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "top")
```

Corporate customers show the lowest rate ratio — they stay the shortest of all customer types, consistent with single-night business trips. Guests on full-board and half-board meal plans stay significantly longer than BB guests: pre-committing to meals signals a structurally longer, holiday-oriented trip. Summer arrivals generate longer stays than all other seasons, driven by leisure travellers with more flexible schedules. Repeated guests stay longer than first-time visitors, likely because familiarity reduces friction and increases the perceived value of extending a stay.

## Post-Hoc Contrasts: Which Segments Differ?

```{r poisson-glht, dependson = "quasipoisson-refit", fig.cap = "Figure 11: Tukey-corrected pairwise contrasts between market segments.", fig.height = 6}
glht_pois <- glht(model_pois, linfct = mcp(MarketSegment = "Tukey"))

glht_df <- as.data.frame(confint(glht_pois)$confint) %>%
  rownames_to_column("Contrast") %>%
  rename(estimate = Estimate, lower = lwr, upper = upr) %>%
  mutate(
    rr          = exp(estimate),
    rr_low      = exp(lower),
    rr_high     = exp(upper),
    significant = !(rr_low <= 1 & rr_high >= 1)
  )

ggplot(glht_df, aes(x = rr, y = reorder(Contrast, rr), colour = significant)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = rr_low, xmax = rr_high), height = 0.4, linewidth = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "darkred", linewidth = 0.8) +
  scale_colour_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey60"),
                      labels = c("TRUE" = "Significant (Tukey)", "FALSE" = "Not significant"),
                      name = NULL) +
  scale_x_continuous(labels = function(x) paste0(x, "×")) +
  labs(title = "Pairwise Market Segment Contrasts: Stay Length",
       subtitle = "Tukey-corrected; rate ratio > 1× means first segment stays longer",
       x = "Rate Ratio (exp[contrast])", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top")
```

The most robust distinctions are between Corporate bookings and every leisure-oriented segment — this gap is large, precise, and survives the multiple-comparison correction. Differences between Online TA and Direct are smaller and more uncertain. For a revenue manager, the actionable finding is the Corporate vs. leisure gap: stay-extension offers directed at Online TA leisure guests in Summer have the strongest evidence base; targeting Corporate guests for the same promotion is unlikely to move the needle.

## Test Set Performance and Business Takeaway

```{r poisson-test-performance, dependson = "quasipoisson-refit"}
pred_pois <- predict(model_pois, newdata = test_pois, type = "response")
mae_pois  <- mean(abs(pred_pois - test_pois$total_stays))
rmse_pois <- sqrt(mean((pred_pois - test_pois$total_stays)^2))
cat(sprintf("Mean Absolute Error: %.2f nights\n", mae_pois))
cat(sprintf("RMSE:                %.2f nights\n", rmse_pois))
```

The quasi-Poisson model predicts stay length to within approximately `r round(mae_pois, 1)` nights on unseen data — useful precision for segment-level planning. A discounted third or fourth night offered at the point of booking for stays above seven days could materially shift average length-of-stay upward for Summer Online TA leisure bookings, generating incremental revenue with minimal additional operational cost.

---

# Model 3: Binomial GLM — Which Bookings Will Cancel?

*Section lead: Lorena Zovkic*

## The Business Problem

Every cancellation is a double loss: the room sits empty, and the revenue that could have been earned from another guest is gone. The earlier a high-risk booking can be identified — at the moment of reservation, before any room is allocated — the more options the hotel has: a retention communication, an adjusted overbooking allowance, a targeted offer. This model answers a precise operational question: given only the information available on day one, how likely is it that this guest will cancel?

## Why Logistic Regression?

`IsCanceled` takes exactly two values. The Binomial GLM models the log-odds of cancellation as a linear function of the predictors, guaranteeing that every prediction stays within the valid probability range [0, 1]. The coefficients it produces are odds ratios — a directly actionable quantity that tells a revenue manager not just whether a predictor matters, but by how much and in which direction. Only information available at the time of booking is used as predictors.

The model carries one important assumption: each predictor contributes additively and linearly to the log-odds of cancellation. For most categorical predictors this is entirely reasonable. For lead time, Figure EDA-5 already showed the true relationship may curve. The Binomial GLM is fitted here as a transparent, interpretable baseline; Model 4 directly tests where linearity breaks down.

## Data Preparation

```{r binomial-prep, cache = TRUE}
hotel_bin <- hotel_data %>%
  filter(!as.character(MarketSegment) %in% c("Undefined", "Complementary"))
hotel_bin$MarketSegment <- droplevels(hotel_bin$MarketSegment)

# Shared colour palette for all cancellation model plots (Models 3–6).
# Defined here once; used in every subsequent plot chunk.
clr <- c(
  primary   = "#2E5FA3",
  secondary = "#E05A2B",
  neutral   = "#6B7280",
  success   = "#1E7A34",
  warning_c = "#B45309"
)

set.seed(2026)
idx_bin   <- createDataPartition(hotel_bin$IsCanceled, p = 0.8, list = FALSE)
train_bin <- hotel_bin[idx_bin, ]
test_bin  <- hotel_bin[-idx_bin, ]

cat(sprintf(
  "Training: %d | Test: %d | Cancellation rate (train): %.1f%%\n",
  nrow(train_bin), nrow(test_bin),
  mean(train_bin$IsCanceled) * 100
))
```

The 80/20 split is stratified on `IsCanceled` to preserve the 41.7% cancellation rate in both sets. The same `train_bin` and `test_bin` objects are reused identically for Models 4, 5, and 6 — this is the foundation for a fair four-way comparison.

## Exploratory Analysis

```{r binomial-eda, fig.cap = "Figure 12: Cancellation rate by deposit type (left) and cancellation probability by lead time (right).", fig.width = 10, fig.height = 4}
p_dep <- train_bin %>%
  group_by(DepositType) %>%
  summarise(rate = mean(IsCanceled), n = n(), .groups = "drop") %>%
  ggplot(aes(x = reorder(DepositType, rate), y = rate, fill = DepositType)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(scales::percent(rate, accuracy = 0.1),
                               "\n(n = ", scales::comma(n), ")")),
            hjust = -0.08, size = 3.2, colour = clr["neutral"]) +
  coord_flip() +
  scale_fill_manual(values = c("No Deposit" = clr["primary"],
                               "Non Refund" = clr["secondary"],
                               "Refundable" = clr["warning_c"])) +
  scale_y_continuous(labels = scales::percent_format(),
                     limits = c(0, 1.3), expand = c(0, 0)) +
  labs(title = "Cancellation Rate by Deposit Type",
       subtitle = "Non Refund: 99.8% — recording artefact, not causal",
       x = NULL, y = "Cancellation Rate") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))

p_lead <- train_bin %>%
  slice_sample(n = 10000) %>%
  ggplot(aes(x = LeadTime, y = IsCanceled)) +
  geom_smooth(method = "loess", span = 0.35,
              colour = clr["primary"],
              fill = adjustcolor(clr["primary"], 0.15), linewidth = 1.3) +
  geom_hline(yintercept = mean(train_bin$IsCanceled),
             linetype = "dashed", colour = clr["secondary"], linewidth = 0.7) +
  annotate("text", x = 390, y = mean(train_bin$IsCanceled) + 0.05,
           label = paste0("Overall rate: ",
                          scales::percent(mean(train_bin$IsCanceled), accuracy = 0.1)),
           colour = clr["secondary"], size = 3, fontface = "italic") +
  scale_x_continuous(limits = c(0, 500)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.92)) +
  labs(title = "Cancellation Probability by Lead Time",
       subtitle = "LOESS smoother — risk rises steeply, then plateaus after ~150 days",
       x = "Lead Time (days before arrival)", y = "P(Cancellation)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))

p_dep + p_lead
```

The left panel confirms the deposit-type artefact described in the EDA. The right panel reproduces — now on the training set only — the nonlinear lead-time pattern from Figure EDA-5. Both findings were anticipated; they are included here to confirm that the training split preserved the patterns of the full dataset.

## Model Fitting

```{r binomial-fit, cache = TRUE, dependson = "binomial-prep"}
model_binomial <- glm(
  IsCanceled ~ LeadTime + DepositType + MarketSegment +
    PreviousCancellations + TotalOfSpecialRequests +
    IsRepeatedGuest + BookingChanges + DaysInWaitingList +
    CustomerType + is_via_agent + arrival_season +
    avg_temp_c + precipitation_mm,
  data   = train_bin,
  family = binomial
)

null_dev <- model_binomial$null.deviance
res_dev  <- deviance(model_binomial)
mcf_r2   <- round(1 - res_dev / null_dev, 3)

cat(sprintf("Null deviance:      %.0f\n", null_dev))
cat(sprintf("Residual deviance:  %.0f  (%.0f%% reduction)\n",
            res_dev, (1 - res_dev / null_dev) * 100))
cat(sprintf("McFadden pseudo-R²: %.3f\n", mcf_r2))
```

## Formal Predictor Testing with drop1()

```{r binomial-drop1, dependson = "binomial-fit", fig.cap = "Figure 13: Chi-square statistics from drop1(). Each bar quantifies how much model fit deteriorates when that predictor is removed entirely.", fig.height = 5}
drop1_bin <- drop1(model_binomial, test = "Chisq")

drop1_bin_df <- as.data.frame(drop1_bin) %>%
  rownames_to_column("Predictor") %>%
  filter(Predictor != "<none>") %>%
  mutate(Predictor = dplyr::recode(Predictor,
    "LeadTime"               = "Lead Time",
    "DepositType"            = "Deposit Type",
    "MarketSegment"          = "Market Segment",
    "PreviousCancellations"  = "Previous Cancellations",
    "TotalOfSpecialRequests" = "Special Requests",
    "IsRepeatedGuest"        = "Repeated Guest",
    "BookingChanges"         = "Booking Changes",
    "DaysInWaitingList"      = "Days on Waiting List",
    "CustomerType"           = "Customer Type",
    "is_via_agent"           = "Booked via Agent",
    "arrival_season"         = "Arrival Season",
    "avg_temp_c"             = "Avg. Temperature",
    "precipitation_mm"       = "Precipitation"
  ))

crit_chi <- qchisq(0.95, df = 1)

ggplot(drop1_bin_df, aes(x = LRT, y = reorder(Predictor, LRT))) +
  geom_col(aes(fill = LRT > crit_chi), width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = ifelse(LRT > 1000, paste0(round(LRT / 1000, 1), "k"),
                               as.character(round(LRT, 0)))),
            hjust = -0.15, size = 3.1, colour = clr["neutral"]) +
  geom_vline(xintercept = crit_chi, linetype = "dashed",
             colour = clr["secondary"], linewidth = 0.9) +
  annotate("text", x = crit_chi + 40, y = 0.7,
           label = paste0("α = 0.05\n(χ² = ", round(crit_chi, 1), ")"),
           colour = clr["secondary"], size = 3, hjust = 0) +
  scale_fill_manual(values = c("TRUE" = clr["primary"], "FALSE" = clr["neutral"])) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(title = "Predictor Contribution: drop1() Chi-square Test",
       subtitle = "All predictors contribute significantly — none are redundant",
       x = "Chi-square Statistic (LRT)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))
```

Every predictor clears the significance threshold. Previous cancellations and deposit type cause the largest model deterioration when removed, confirming that booking history and payment commitment are the strongest cancellation signals. Weather variables (temperature, precipitation) contribute modestly but remain statistically justified — they add marginal signal beyond arrival season at the cost of two parameters.

## Coefficient Interpretation: Odds Ratios

```{r binomial-forest, dependson = "binomial-fit", fig.cap = "Figure 14: Odds ratios with 95% confidence intervals on a log scale. Values right of the dashed line increase cancellation odds; values left reduce them. Non Refund deposit is excluded — its OR approaches infinity and reflects a recording artefact.", fig.height = 7.5, fig.width = 9}
bin_tidy <- tidy(model_binomial, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)", !grepl("Non.Refund", term)) %>%
  mutate(
    significant = p.value < 0.05,
    direction   = ifelse(estimate > 1, "Increases risk", "Reduces risk"),
    term_clean  = term %>%
      str_replace("MarketSegment",         "Segment: ") %>%
      str_replace("CustomerType",          "Customer: ") %>%
      str_replace("DepositTypeRefundable", "Deposit: Refundable") %>%
      str_replace("arrival_season",        "Season: ") %>%
      str_replace("LeadTime",              "Lead Time (per day)") %>%
      str_replace("PreviousCancellations", "Previous Cancellations") %>%
      str_replace("TotalOfSpecialRequests","Special Requests") %>%
      str_replace("IsRepeatedGuest",       "Repeated Guest") %>%
      str_replace("BookingChanges",        "Booking Changes") %>%
      str_replace("DaysInWaitingList",     "Days on Waiting List") %>%
      str_replace("is_via_agent",          "Booked via Agent") %>%
      str_replace("avg_temp_c",            "Avg. Temperature") %>%
      str_replace("precipitation_mm",      "Precipitation")
  )

ggplot(bin_tidy, aes(x = estimate, y = reorder(term_clean, estimate))) +
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = clr["neutral"], linewidth = 0.8) +
  geom_segment(aes(x = conf.low, xend = conf.high,
                   yend = reorder(term_clean, estimate)),
               colour = "grey80", linewidth = 1.2) +
  geom_point(aes(colour = significant, shape = direction), size = 3.5) +
  scale_colour_manual(values = c("TRUE" = clr["primary"], "FALSE" = "grey55"),
                      labels = c("TRUE" = "p < 0.05", "FALSE" = "p ≥ 0.05"),
                      name = "Significance") +
  scale_shape_manual(values = c("Increases risk" = 17, "Reduces risk" = 19),
                     name = "Direction") +
  scale_x_log10(labels = function(x) paste0(x, "×"),
                breaks = c(0.2, 0.4, 0.6, 0.8, 1, 1.5, 2, 3, 5, 8)) +
  labs(title = "Binomial GLM — Odds Ratios for Cancellation",
       subtitle = "Non Refund deposit omitted (OR ≈ ∞ — recording artefact)",
       x = "Odds Ratio [exp(β)] — log scale", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))
```

Five findings stand out for operational interpretation. **Lead time:** the per-day OR of `r round(exp(coef(model_binomial)["LeadTime"]), 4)` compounds to `r round(exp(coef(model_binomial)["LeadTime"] * 30), 2)`× per 30 days — a booking made six months ahead carries roughly `r round(exp(coef(model_binomial)["LeadTime"]*180)/exp(coef(model_binomial)["LeadTime"]*7),1)`× higher odds than one made a week out. **Special requests:** each additional request reduces cancellation odds by `r round((1 - exp(coef(model_binomial)["TotalOfSpecialRequests"]))*100, 0)`% — the strongest protective factor, reflecting psychological commitment to the trip. **Previous cancellations:** each prior cancellation multiplies current odds by `r round(exp(coef(model_binomial)["PreviousCancellations"]), 2)`× — booking history is highly predictive and already sits in the hotel's reservation system. **Booking changes:** counterintuitively, each modification *reduces* odds by `r round((1-exp(coef(model_binomial)["BookingChanges"]))*100,0)`% — a guest actively managing their booking plans to show up. **Repeated guests:** returning visitors are `r round((1-exp(coef(model_binomial)["IsRepeatedGuest"]))*100,0)`% less likely to cancel, reflecting established commitment to the property.

## Marginal Effect of Lead Time

```{r binomial-marginal, dependson = "binomial-fit", fig.cap = "Figure 15: Predicted cancellation probability by lead time for a typical reference booking. The perfectly straight line is a structural consequence of the GLM's linearity assumption — Model 4 tests whether this is accurate.", fig.width = 8, fig.height = 4.5}
lead_seq_bin <- seq(0, 400, by = 2)

newdata_bin <- data.frame(
  LeadTime               = lead_seq_bin,
  DepositType            = factor("No Deposit", levels = levels(train_bin$DepositType)),
  MarketSegment          = factor("Online TA",  levels = levels(train_bin$MarketSegment)),
  PreviousCancellations  = 0,
  TotalOfSpecialRequests = 1,
  IsRepeatedGuest        = 0L,
  BookingChanges         = 0,
  DaysInWaitingList      = 0,
  CustomerType           = factor("Transient",  levels = levels(train_bin$CustomerType)),
  is_via_agent           = 1L,
  arrival_season         = factor("Summer",     levels = levels(train_bin$arrival_season)),
  avg_temp_c             = mean(train_bin$avg_temp_c,       na.rm = TRUE),
  precipitation_mm       = mean(train_bin$precipitation_mm, na.rm = TRUE)
)

preds_bin            <- predict(model_binomial, newdata = newdata_bin,
                                type = "link", se.fit = TRUE)
newdata_bin$prob     <- plogis(preds_bin$fit)
newdata_bin$prob_lwr <- plogis(preds_bin$fit - 1.96 * preds_bin$se.fit)
newdata_bin$prob_upr <- plogis(preds_bin$fit + 1.96 * preds_bin$se.fit)

ggplot(newdata_bin, aes(x = LeadTime)) +
  geom_ribbon(aes(ymin = prob_lwr, ymax = prob_upr),
              fill = clr["primary"], alpha = 0.15) +
  geom_line(aes(y = prob), colour = clr["primary"], linewidth = 1.4) +
  geom_hline(yintercept = mean(train_bin$IsCanceled),
             linetype = "dashed", colour = clr["secondary"], linewidth = 0.7) +
  annotate("text", x = 300, y = mean(train_bin$IsCanceled) + 0.04,
           label = paste0("Overall cancellation rate: ",
                          scales::percent(mean(train_bin$IsCanceled), accuracy = 0.1)),
           colour = clr["secondary"], size = 3, fontface = "italic") +
  annotate("text", x = 355, y = 0.80,
           label = "The GLM assumes\na straight-line increase.\nModel 4 tests this.",
           colour = clr["neutral"], size = 3, fontface = "italic", hjust = 0.5) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.95)) +
  scale_x_continuous(limits = c(0, 400)) +
  labs(title = "Marginal Effect of Lead Time on Cancellation Probability",
       subtitle = "Reference: Online TA · Transient · No Deposit · Summer · 1 special request",
       x = "Lead Time (days before arrival)", y = "Predicted Cancellation Probability") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))
```

## Test Set Performance

```{r binomial-roc, dependson = "binomial-fit", fig.cap = "Figure 16: ROC curve for the Binomial GLM on the held-out test set.", fig.width = 6, fig.height = 5.5}
pred_probs_bin <- predict(model_binomial, newdata = test_bin, type = "response")
roc_bin        <- roc(test_bin$IsCanceled, pred_probs_bin, quiet = TRUE)

roc_df_bin <- data.frame(fpr = 1 - roc_bin$specificities,
                          tpr = roc_bin$sensitivities)

ggplot(roc_df_bin, aes(x = fpr, y = tpr)) +
  geom_ribbon(aes(ymin = 0, ymax = tpr), fill = clr["primary"], alpha = 0.10) +
  geom_line(colour = clr["primary"], linewidth = 1.3) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey65", linewidth = 0.7) +
  annotate("text", x = 0.58, y = 0.30,
           label = paste0("AUC = ", round(auc(roc_bin), 3)),
           colour = clr["primary"], size = 5.5, fontface = "bold") +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(title = "ROC Curve: Binomial GLM",
       subtitle = "Held-out test set — not seen during model training",
       x = "False Positive Rate (1 − Specificity)",
       y = "True Positive Rate (Sensitivity)") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))
```

```{r binomial-confmat, dependson = "binomial-roc"}
pred_class_bin <- factor(ifelse(pred_probs_bin > 0.5, 1, 0))
cm_bin <- caret::confusionMatrix(pred_class_bin, factor(test_bin$IsCanceled),
                                  positive = "1")
cm_bin
```

## Business Takeaway

An AUC of `r round(auc(roc_bin), 3)` on unseen data means the model correctly ranks a randomly drawn cancellation above a randomly drawn non-cancellation `r round(auc(roc_bin)*100, 1)`% of the time — well above the 50% baseline of random guessing. For a revenue manager, three findings translate directly into action: any booking combining a lead time above 100 days, zero special requests, and at least one prior cancellation falls into the highest-risk tier and should trigger proactive retention; the protective effect of special requests suggests prompting guests to submit room or meal preferences within 48 hours of booking; and the counterintuitive booking-changes finding suggests that making modifications easy rather than friction-heavy reduces overall cancellation risk.

---

# Model 4: Generalised Additive Model — Do Effects Follow a Straight Line?

*Section lead: Lorena Zovkic*

## The Question This Model Answers

The Binomial GLM is transparent and actionable — but it forces every continuous predictor's relationship with cancellation to be a straight line on the log-odds scale. We know from Figure EDA-5 that lead time does not behave this way. The GAM keeps the logistic regression framework but replaces linear terms for selected continuous predictors with flexible smooth functions that let the data determine the shape. Three questions:

1. Are the nonlinear patterns in the EDA statistically real, or smoothing artefacts?
2. Does correcting for them improve out-of-sample prediction?
3. Where exactly does the Binomial GLM go wrong — and what are the operational consequences?

## Why a GAM?

The GAM occupies a deliberate position in the modelling ladder (Table 0). It is more flexible than the Binomial GLM (drops linearity for continuous terms) but more constrained than the SVM and Neural Network (retains additivity — each smooth still operates independently). This makes it interpretable: the smooth effect of lead time can be plotted directly and explained to a revenue manager. The complexity of each smooth is determined automatically by penalised regression (REML) — the model uses only as much flexibility as the data justifies. The effective degrees of freedom (EDF) summarises this: EDF = 1 means the smooth collapsed to a straight line (the GLM assumption was correct); EDF >> 1 means genuine nonlinearity was detected.

## Model Fitting

```{r gam-fit, cache = TRUE, dependson = "binomial-prep"}
library(mgcv)

model_gam <- gam(
  IsCanceled ~ s(LeadTime) + s(ADR) + s(DaysInWaitingList) +
    DepositType + MarketSegment + PreviousCancellations +
    TotalOfSpecialRequests + IsRepeatedGuest +
    CustomerType + arrival_season +
    avg_temp_c + precipitation_mm,
  data   = train_bin,
  family = binomial,
  method = "REML"
)

gam_summ <- summary(model_gam)

smooth_table <- as.data.frame(gam_summ$s.table) %>%
  rownames_to_column("Term") %>%
  mutate(
    Term = dplyr::recode(Term,
      "s(LeadTime)"          = "s(Lead Time)",
      "s(ADR)"               = "s(ADR — room price)",
      "s(DaysInWaitingList)" = "s(Days on Waiting List)"
    ),
    Verdict = case_when(
      edf < 1.5 ~ "Approximately linear — GLM assumption holds",
      edf < 3.0 ~ "Mildly nonlinear — GLM slightly misrepresents this",
      TRUE      ~ "Clearly nonlinear — GLM was meaningfully wrong here"
    )
  )

kable(smooth_table %>% dplyr::select(Term, edf, `p-value`, Verdict),
      digits = c(0, 2, 4, 0),
      col.names = c("Smooth Term", "EDF", "p-value", "Verdict"),
      caption = "Table 4-1: EDF for the three smooth terms. EDF = 1 is a straight line; higher values indicate genuine nonlinearity that the Binomial GLM cannot represent.") %>%
  kable_styling(full_width = TRUE, font_size = 11) %>%
  row_spec(which(smooth_table$edf > 2), bold = TRUE, background = "#EFF6FF")

cat(sprintf("\nDeviance explained: %.1f%%\n", gam_summ$dev.expl * 100))
cat(sprintf("AIC:                %.1f\n",    AIC(model_gam)))
```

## What Shape Does Each Relationship Take?

```{r gam-smooths, dependson = "gam-fit", fig.cap = "Figure 17: Estimated smooth effects from the GAM. Each panel isolates one predictor's contribution to the log-odds of cancellation after controlling for all others. The EDF in each title quantifies departure from linearity.", fig.width = 12, fig.height = 4.5}
edf_lead <- round(gam_summ$s.table["s(LeadTime)",          "edf"], 1)
edf_adr  <- round(gam_summ$s.table["s(ADR)",               "edf"], 1)
edf_wl   <- round(gam_summ$s.table["s(DaysInWaitingList)", "edf"], 1)

# Extract smooth effects safely via predict(type = "terms").
# This approach builds a grid that varies one predictor while holding all
# others at their training-set means / modal values, avoiding the fragility
# of directly referencing model$model[1, ].
make_smooth_df <- function(model, var_name, x_label, n = 300) {
  col_data <- model$model[[var_name]]
  x_seq    <- seq(quantile(col_data, 0.01, na.rm = TRUE),
                  quantile(col_data, 0.99, na.rm = TRUE),
                  length.out = n)

  # Build a reference row using training-set means / modal factor levels
  ref_row <- as.data.frame(lapply(model$model, function(col) {
    if (is.numeric(col)) mean(col, na.rm = TRUE)
    else if (is.factor(col)) levels(col)[which.max(tabulate(col))]
    else col[1]
  }))
  ref_row <- ref_row[rep(1, n), ]
  ref_row[[var_name]] <- x_seq

  pt <- predict(model, newdata = ref_row, type = "terms", se.fit = TRUE)
  col_idx <- which(grepl(paste0("(", var_name, ")"), colnames(pt$fit), fixed = TRUE))[1]
  data.frame(x = x_seq, fit = as.numeric(pt$fit[, col_idx]),
             se = as.numeric(pt$se.fit[, col_idx]), x_lab = x_label)
}

d_lead <- make_smooth_df(model_gam, "LeadTime", "Lead Time (days before arrival)")
d_adr  <- make_smooth_df(model_gam, "ADR",      "Average Daily Rate — ADR (€/night)")
d_wl   <- make_smooth_df(model_gam, "DaysInWaitingList", "Days on Waiting List")

plot_smooth <- function(d, title_str, edf_val) {
  ggplot(d, aes(x = x)) +
    geom_ribbon(aes(ymin = fit - 1.96 * se, ymax = fit + 1.96 * se),
                fill = clr["primary"], alpha = 0.15) +
    geom_line(aes(y = fit), colour = clr["primary"], linewidth = 1.3) +
    geom_hline(yintercept = 0, linetype = "dashed",
               colour = clr["neutral"], linewidth = 0.7) +
    labs(title = paste0(title_str, "  (EDF = ", edf_val, ")"),
         x = unique(d$x_lab), y = "Partial effect (log-odds)") +
    theme_minimal(base_size = 10) +
    theme(plot.title = element_text(face = "bold", size = 10),
          panel.grid.minor = element_blank())
}

plot_smooth(d_lead, "s(Lead Time)", edf_lead) +
plot_smooth(d_adr,  "s(ADR)",      edf_adr) +
plot_smooth(d_wl,   "s(Days on Waiting List)", edf_wl)
```

Three panels, three distinct findings. **s(Lead Time) — EDF = `r edf_lead`:** The smooth confirms and quantifies the EDA finding. Risk rises steeply to around 100–150 days then clearly flattens. The GLM's straight line assigns ever-increasing risk past the plateau and meaningfully misrepresents both committed long-planners (overscored) and the steepest early-rise zone (underscored). **s(ADR) — EDF = `r edf_adr`:** Room price has a non-monotone relationship the GLM cannot capture — mid-range priced rooms carry the highest cancellation risk, while very cheap and very expensive rooms show lower risk. Price-sensitive guests use what they paid for; luxury bookings accompany specific purposeful trips; mid-range rooms attract the widest mix of opportunistic bookers. **s(Days on Waiting List) — EDF = `r edf_wl`:** The GAM chose not to use its extra flexibility here — a straight line fits as well as a curve. This is a valuable null result: not every predictor needs a smooth, and the model correctly identifies where the GLM's linearity assumption holds.

## Does the Flexibility Improve Prediction?

```{r gam-comparison, dependson = c("gam-fit", "binomial-fit")}
aic_glm  <- AIC(model_binomial)
aic_gam  <- AIC(model_gam)
aic_diff <- round(aic_glm - aic_gam, 0)

cat(sprintf("AIC — Binomial GLM: %.1f\n", aic_glm))
cat(sprintf("AIC — GAM:          %.1f\n", aic_gam))
cat(sprintf("AIC improvement:    %.0f points\n", aic_diff))
```

```{r gam-roc, dependson = c("gam-fit", "binomial-roc"), fig.cap = "Figure 18: ROC curves for the Binomial GLM and GAM on the same held-out test set. The shaded rectangle marks the operating range (FPR 10–35%) most relevant for a hotel risk-scoring system.", fig.width = 7, fig.height = 5.5}
gam_probs <- predict(model_gam, newdata = test_bin, type = "response")
roc_gam   <- roc(test_bin$IsCanceled, gam_probs, quiet = TRUE)

glm_lbl <- paste0("Binomial GLM  (AUC = ", round(auc(roc_bin), 3), ")")
gam_lbl <- paste0("GAM           (AUC = ", round(auc(roc_gam), 3), ")")

bind_rows(
  data.frame(fpr = 1 - roc_bin$specificities,
             tpr = roc_bin$sensitivities, Model = glm_lbl),
  data.frame(fpr = 1 - roc_gam$specificities,
             tpr = roc_gam$sensitivities, Model = gam_lbl)
) %>%
  ggplot(aes(x = fpr, y = tpr, colour = Model)) +
  geom_line(linewidth = 1.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey65", linewidth = 0.7) +
  annotate("rect", xmin = 0.05, xmax = 0.35, ymin = 0.55, ymax = 1.0,
           fill = "grey96", colour = "grey85", linewidth = 0.3) +
  annotate("text", x = 0.20, y = 0.61,
           label = "Operationally relevant\nregion for risk scoring",
           colour = clr["neutral"], size = 2.8, fontface = "italic") +
  scale_colour_manual(values = setNames(c(clr["primary"], clr["secondary"]),
                                        c(glm_lbl, gam_lbl))) +
  scale_x_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(title = "ROC Comparison: Binomial GLM vs. GAM",
       subtitle = "Same predictors, same data — difference reflects smooth terms only",
       x = "False Positive Rate", y = "True Positive Rate", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))
```

```{r gam-confmat, dependson = "gam-roc"}
pred_class_gam <- factor(ifelse(gam_probs > 0.5, 1, 0))
cm_gam <- caret::confusionMatrix(pred_class_gam, factor(test_bin$IsCanceled),
                                  positive = "1")
cm_gam
```

The GAM achieves an AUC of `r round(auc(roc_gam), 3)` versus `r round(auc(roc_bin), 3)` for the GLM — a gain of `r round(auc(roc_gam) - auc(roc_bin), 3)` points on unseen data. The AIC drops by `r aic_diff` points, confirming the improvement is genuine and not overfitting.

## Where Exactly Do GLM and GAM Disagree?

```{r gam-marginal, dependson = c("gam-fit", "binomial-fit"), fig.cap = "Figure 19: Predicted cancellation probability from the GAM (solid) and GLM (dashed) for the same reference booking. Annotations identify the two zones of meaningful operational disagreement.", fig.width = 8.5, fig.height = 4.5}
newdata_gam <- data.frame(
  LeadTime               = lead_seq_bin,
  ADR                    = median(train_bin$ADR[train_bin$ADR > 0], na.rm = TRUE),
  DaysInWaitingList      = 0,
  DepositType            = factor("No Deposit", levels = levels(train_bin$DepositType)),
  MarketSegment          = factor("Online TA",  levels = levels(train_bin$MarketSegment)),
  PreviousCancellations  = 0,
  TotalOfSpecialRequests = 1,
  IsRepeatedGuest        = 0L,
  CustomerType           = factor("Transient",  levels = levels(train_bin$CustomerType)),
  arrival_season         = factor("Summer",     levels = levels(train_bin$arrival_season)),
  avg_temp_c             = mean(train_bin$avg_temp_c,       na.rm = TRUE),
  precipitation_mm       = mean(train_bin$precipitation_mm, na.rm = TRUE)
)

gam_link   <- predict(model_gam, newdata = newdata_gam, type = "link", se.fit = TRUE)
gam_margin <- data.frame(LeadTime = lead_seq_bin,
                         prob     = plogis(gam_link$fit),
                         prob_lwr = plogis(gam_link$fit - 1.96 * gam_link$se.fit),
                         prob_upr = plogis(gam_link$fit + 1.96 * gam_link$se.fit))

ggplot() +
  geom_ribbon(data = gam_margin,
              aes(x = LeadTime, ymin = prob_lwr, ymax = prob_upr),
              fill = clr["secondary"], alpha = 0.12) +
  geom_line(data = gam_margin,
            aes(x = LeadTime, y = prob, colour = gam_lbl), linewidth = 1.4) +
  geom_line(data = data.frame(LeadTime = lead_seq_bin, prob = newdata_bin$prob),
            aes(x = LeadTime, y = prob, colour = glm_lbl),
            linewidth = 1.1, linetype = "dashed") +
  annotate("segment", x = 150, xend = 150, y = 0.05, yend = 0.87,
           linetype = "dotted", colour = clr["neutral"], linewidth = 0.8) +
  annotate("text", x = 154, y = 0.84,
           label = "GAM plateau\nbegins ~150 days",
           colour = clr["neutral"], size = 2.9, fontface = "italic", hjust = 0) +
  annotate("label", x = 265, y = 0.52,
           label = "GLM overestimates\nrisk in this zone",
           colour = clr["primary"], size = 2.9, fill = "white", label.size = 0.25) +
  annotate("label", x = 22, y = 0.27,
           label = "GLM underestimates\nrisk here",
           colour = clr["primary"], size = 2.9, fill = "white", label.size = 0.25) +
  scale_colour_manual(values = setNames(c(clr["secondary"], clr["primary"]),
                                        c(gam_lbl, glm_lbl))) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_x_continuous(limits = c(0, 400)) +
  labs(title = "Where Do GLM and GAM Disagree?",
       subtitle = "Reference: Online TA · Transient · No Deposit · Summer · median ADR · 1 special request",
       x = "Lead Time (days before arrival)",
       y = "Predicted Cancellation Probability", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = clr["neutral"], size = 9))
```

## Business Takeaway

The GAM is strictly better than the Binomial GLM for cancellation prediction — higher AUC on held-out data, lower AIC, and a more accurate representation of how lead time and ADR actually relate to cancellation risk. It achieves this without sacrificing interpretability: each smooth is a curve that can be shown to a revenue manager. Two operational implications follow from the smooth terms: lead-time risk plateaus after 150 days (treat bookings beyond that threshold as equivalent risk, free up overbooking capacity), and mid-range ADR rooms carry disproportionate risk (consider targeted deposit requirements for this price band). For the scoring layer, the GAM is the right model. For explaining a flagged booking to a commercial stakeholder, the GLM's odds ratios remain the cleaner communication tool — the two models are complementary.

---

# Model 5: Support Vector Machine

*Section lead: Petronela Tabalae*

## Research Question

> **Can an SVM improve cancellation classification beyond the GLM and GAM, and does the gain come from the kernel's ability to represent feature interactions?**

Both the GLM and GAM are additive: each predictor's contribution is independent of all others. Yet booking risk may not decompose this way. A long lead time may be routine for a Corporate guest and alarming for a leisure Online TA guest; a non-refundable deposit may eliminate risk for one customer type and merely defer it for another. If interactions of this kind shape cancellation, only models capable of representing nonlinear joint decision boundaries can capture them. The SVM with an RBF kernel is the canonical tool for this — and comparing it to a linear-kernel SVM isolates exactly how much of any gain comes from interactions versus other differences in how SVMs are trained.

## Data Preparation and Subsampling Rationale

```{r svm-setup, message = FALSE, warning = FALSE}
library(kernlab)
library(PRROC)
library(parallel)

set.seed(2026)
sub_idx       <- createDataPartition(train_bin$IsCanceled,
                                     p = 5000 / nrow(train_bin), list = FALSE)
svm_train_sub <- train_bin[sub_idx, ]

cat(sprintf("Full training set: %d | SVM subsample: %d | Test set: %d\n",
            nrow(train_bin), nrow(svm_train_sub), nrow(test_bin)))
cat(sprintf("Cancellation rate — full: %.3f | subsample: %.3f | test: %.3f\n",
            mean(train_bin$IsCanceled), mean(svm_train_sub$IsCanceled),
            mean(test_bin$IsCanceled)))
```

Kernel SVM training complexity grows roughly as O(n²) — fitting on the full `r format(nrow(train_bin), big.mark=",")` training rows with cross-validation is computationally infeasible on standard hardware. We work on a stratified 5,000-row subsample that preserves the 41.7% cancellation rate exactly. This is documented standard practice for kernel SVMs at this data scale. Critically, **the test set evaluation uses the full held-out test set** — reported test performance is directly comparable to Models 3 and 4.

```{r svm-formula-and-matrix}
svm_formula <- IsCanceled ~ LeadTime + DepositType + MarketSegment +
  PreviousCancellations + TotalOfSpecialRequests + IsRepeatedGuest +
  BookingChanges + DaysInWaitingList + CustomerType + is_via_agent +
  arrival_season + avg_temp_c + precipitation_mm

make_xy <- function(df, scaler = NULL) {
  X <- model.matrix(svm_formula, data = df)[, -1, drop = FALSE]
  if (is.null(scaler)) {
    scaler <- list(center = colMeans(X), scale = apply(X, 2, sd))
    scaler$scale[scaler$scale == 0] <- 1
  }
  X <- scale(X, center = scaler$center, scale = scaler$scale)
  list(X = X,
       y = factor(df$IsCanceled, levels = c(0, 1), labels = c("Stay", "Cancel")),
       scaler = scaler)
}

train_xy <- make_xy(svm_train_sub)
test_xy  <- make_xy(test_bin, scaler = train_xy$scaler)
```

## Cross-Validation for Kernel and Hyperparameter Selection

```{r svm-cv-tuning, cache = TRUE}
set.seed(2026)
folds <- createFolds(svm_train_sub$IsCanceled, k = 5, returnTrain = FALSE)

n_cores <- max(1, min(5, parallel::detectCores() - 1))
cl <- parallel::makeCluster(n_cores)
parallel::clusterEvalQ(cl, { library(kernlab); library(pROC) })
parallel::clusterExport(cl, c("train_xy"))

fit_one_fold <- function(val_idx, C, sigma, kernel) {
  Xtr <- train_xy$X[-val_idx, ]; ytr <- train_xy$y[-val_idx]
  Xva <- train_xy$X[ val_idx, ]; yva <- train_xy$y[ val_idx]
  fit <- if (kernel == "linear") {
    ksvm(Xtr, ytr, type = "C-svc", kernel = "vanilladot", C = C, scaled = FALSE)
  } else {
    ksvm(Xtr, ytr, type = "C-svc", kernel = "rbfdot",
         kpar = list(sigma = sigma), C = C, scaled = FALSE)
  }
  score <- as.numeric(predict(fit, Xva, type = "decision"))
  as.numeric(pROC::auc(yva, score, levels = c("Stay", "Cancel"),
                        direction = "<", quiet = TRUE))
}

cv_auc <- function(C, sigma = NULL, kernel) {
  mean(unlist(parallel::parLapply(cl, folds, fit_one_fold,
                                   C = C, sigma = sigma, kernel = kernel)))
}

linear_grid <- expand.grid(C = c(0.1, 1, 10))
linear_grid$cv_auc <- mapply(function(C) cv_auc(C, kernel = "linear"), linear_grid$C)

rbf_grid <- expand.grid(C = c(0.1, 1, 10), sigma = c(0.01, 0.1, 1))
rbf_grid$cv_auc <- mapply(function(C, s) cv_auc(C, sigma = s, kernel = "rbf"),
                           rbf_grid$C, rbf_grid$sigma)

parallel::stopCluster(cl)
```

```{r svm-cv-display}
kable(linear_grid, digits = 4,
      caption = "Table 2: Linear-kernel SVM — 5-fold CV AUC by cost parameter.") %>%
  kable_styling(full_width = FALSE, font_size = 11)

kable(rbf_grid, digits = 4,
      caption = "Table 3: RBF-kernel SVM — 5-fold CV AUC by (C, σ).") %>%
  kable_styling(full_width = FALSE, font_size = 11)
```

```{r svm-cv-heatmap, fig.cap = "Figure 20: Cross-validation AUC surface for the RBF kernel. The interior optimum confirms the grid range is adequate.", fig.width = 7, fig.height = 4}
ggplot(rbf_grid, aes(x = factor(sigma), y = factor(C), fill = cv_auc)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.3f", cv_auc)), colour = "white", size = 4) +
  scale_fill_viridis_c(option = "D",
                       limits = c(min(rbf_grid$cv_auc), max(rbf_grid$cv_auc))) +
  labs(title = "RBF SVM — Cross-Validated AUC Surface",
       subtitle = "5-fold stratified CV on 5,000-booking training subsample",
       x = expression(sigma~"(kernel width)"),
       y = "C (cost / regularisation strength)", fill = "CV AUC") +
  theme_minimal(base_size = 11)
```

The best linear-kernel configuration achieves CV AUC of `r round(max(linear_grid$cv_auc), 3)`; the best RBF configuration reaches `r round(max(rbf_grid$cv_auc), 3)`. The gap of `r round(max(rbf_grid$cv_auc) - max(linear_grid$cv_auc), 3)` AUC points is the cross-validation-estimated contribution of nonlinear feature interactions within the SVM family.

## Final Models and Test Set Evaluation

```{r svm-final-fit, cache = TRUE}
best_linear <- linear_grid[which.max(linear_grid$cv_auc), ]
best_rbf    <- rbf_grid[which.max(rbf_grid$cv_auc), ]

set.seed(2026)
fit_linear <- ksvm(train_xy$X, train_xy$y, type = "C-svc",
                   kernel = "vanilladot", C = best_linear$C, scaled = FALSE)
set.seed(2026)
fit_rbf <- ksvm(train_xy$X, train_xy$y, type = "C-svc",
                kernel = "rbfdot", kpar = list(sigma = best_rbf$sigma),
                C = best_rbf$C, scaled = FALSE)
```

```{r svm-test-evaluation}
# Decision values are used rather than type = "probabilities" to sidestep
# a known kernlab bug where the linear-kernel prob.model slot is sometimes
# empty. The monotonic logistic transform preserves AUC and PR-AUC exactly.
score_to_prob <- function(s) plogis(as.numeric(scale(s)))

dec_linear  <- as.numeric(predict(fit_linear, test_xy$X, type = "decision"))
dec_rbf     <- as.numeric(predict(fit_rbf,    test_xy$X, type = "decision"))
prob_linear <- score_to_prob(dec_linear)
prob_rbf    <- score_to_prob(dec_rbf)

eval_one <- function(prob, label) {
  pred <- factor(ifelse(prob > 0.5, "Cancel", "Stay"), levels = c("Stay", "Cancel"))
  cm   <- caret::confusionMatrix(pred, test_xy$y, positive = "Cancel")
  rocp <- pROC::roc(test_xy$y, prob, levels = c("Stay", "Cancel"),
                    direction = "<", quiet = TRUE)
  prc  <- PRROC::pr.curve(scores.class0 = prob[test_xy$y == "Cancel"],
                          scores.class1 = prob[test_xy$y == "Stay"], curve = FALSE)
  data.frame(Model    = label,
             Accuracy = unname(cm$overall["Accuracy"]),
             ROC_AUC  = as.numeric(rocp$auc),
             F1       = unname(cm$byClass["F1"]),
             PR_AUC   = prc$auc.integral)
}

svm_results <- rbind(eval_one(prob_linear, "SVM (Linear)"),
                     eval_one(prob_rbf,    "SVM (RBF)"))

kable(svm_results, digits = 4,
      caption = "Table 4: SVM test-set performance on the full held-out test set.") %>%
  kable_styling(full_width = FALSE, font_size = 11)
```

```{r svm-roc-plot, fig.cap = "Figure 21: ROC curves for both SVM kernels alongside the GLM and GAM baselines. All four evaluated on the identical held-out test set.", fig.width = 6, fig.height = 5}
roc_lin <- pROC::roc(test_xy$y, prob_linear, levels = c("Stay", "Cancel"),
                     direction = "<", quiet = TRUE)
roc_rbf <- pROC::roc(test_xy$y, prob_rbf,    levels = c("Stay", "Cancel"),
                     direction = "<", quiet = TRUE)

plot(roc_bin, col = "steelblue", lwd = 2,
     main = "ROC Curves: GLM, GAM, and SVM Kernels")
lines(roc_gam, col = "tomato",    lwd = 2)
lines(roc_lin, col = "grey45",    lwd = 2, lty = 2)
lines(roc_rbf, col = "darkgreen", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright", bty = "n",
       legend = c(paste0("Binomial GLM (AUC = ", round(auc(roc_bin), 3), ")"),
                  paste0("GAM         (AUC = ", round(auc(roc_gam), 3), ")"),
                  paste0("SVM Linear  (AUC = ", round(auc(roc_lin), 3), ")"),
                  paste0("SVM RBF     (AUC = ", round(auc(roc_rbf), 3), ")")),
       col = c("steelblue", "tomato", "grey45", "darkgreen"),
       lwd = 2, lty = c(1, 1, 2, 1))
```

The linear-kernel SVM (grey dashed) sits close to the Binomial GLM — two models that both operate in the original feature space without explicit interaction terms. This convergence confirms that the AUC gap between the linear and RBF SVMs is attributable to the kernel's ability to represent interactions, not to any difference between margin-maximisation and maximum-likelihood as fitting principles.

## Business Takeaway

The RBF SVM achieves a test-set AUC of `r round(auc(roc_rbf), 3)`, versus `r round(auc(roc_lin), 3)` for the linear kernel. The gap of `r round(auc(roc_rbf) - auc(roc_lin), 3)` AUC points is the operational benefit of allowing the model to capture feature interactions. For a hotel deploying a risk-scoring system where prediction accuracy is the primary value, the RBF SVM is the right scoring engine. Where the hotel needs to explain a flagged booking to a commercial stakeholder, the GLM's odds ratios remain the communication tool of choice.

---

# Model 6: Neural Network

*Section lead: Petronela Tabalae*

## Research Question

> **Can a feedforward neural network recover cancellation patterns that additive models miss, and which predictors does it rely on most differently compared with the Binomial GLM?**

The neural network goes one step further than the SVM: hidden units learn their own basis functions from the data, without the analyst specifying any structure. If the cancellation process contains interactions that neither additive structure nor a Gaussian kernel naturally express, the neural network offers the most flexible representation in this report's modelling arsenal.

## Why a Single-Hidden-Layer Network?

We use `nnet::nnet` rather than a deep-learning framework. For tabular data with approximately twenty effective predictors after dummy expansion, a single hidden-layer network trained by BFGS converges in seconds on a CPU, has a small and well-understood hyperparameter space (hidden layer size, weight decay), and is exactly reproducible from a seed. Deeper architectures bring diminishing returns on tabular data of this kind. Unlike the kernel SVM, the neural network's training cost grows linearly with sample size — we fit on the full training set.

## Data Preparation

```{r nnet-setup, message = FALSE, warning = FALSE}
library(nnet)

nn_formula <- IsCanceled ~ LeadTime + DepositType + MarketSegment +
  PreviousCancellations + TotalOfSpecialRequests + IsRepeatedGuest +
  BookingChanges + DaysInWaitingList + CustomerType + is_via_agent +
  arrival_season + avg_temp_c + precipitation_mm

make_nn_xy <- function(df, scaler = NULL) {
  X <- model.matrix(nn_formula, data = df)[, -1, drop = FALSE]
  if (is.null(scaler)) {
    scaler <- list(center = colMeans(X), scale = apply(X, 2, sd))
    scaler$scale[scaler$scale == 0] <- 1
  }
  X <- scale(X, center = scaler$center, scale = scaler$scale)
  list(X = X,
       y = factor(df$IsCanceled, levels = c(0, 1), labels = c("Stay", "Cancel")),
       scaler = scaler)
}

nn_train <- make_nn_xy(train_bin)
nn_test  <- make_nn_xy(test_bin, scaler = nn_train$scaler)

cat(sprintf("Training rows: %d | Test rows: %d | Predictors: %d\n",
            nrow(nn_train$X), nrow(nn_test$X), ncol(nn_train$X)))
```

## Cross-Validation for Architecture Selection

```{r nnet-cv-tuning, cache = TRUE}
set.seed(2026)
nn_folds <- createFolds(train_bin$IsCanceled, k = 5, returnTrain = FALSE)

nn_cores <- max(1, min(5, parallel::detectCores() - 1))
nn_cl <- parallel::makeCluster(nn_cores)
parallel::clusterEvalQ(nn_cl, { library(nnet); library(pROC) })
parallel::clusterExport(nn_cl, c("nn_train"))

fit_one_nn_fold <- function(val_idx, size, decay) {
  Xtr <- nn_train$X[-val_idx, ]; ytr <- nn_train$y[-val_idx]
  Xva <- nn_train$X[ val_idx, ]; yva <- nn_train$y[ val_idx]
  set.seed(2026)
  fit <- nnet(x = Xtr, y = (ytr == "Cancel") * 1L,
              size = size, decay = decay, maxit = 200,
              trace = FALSE, MaxNWts = 5000, linout = FALSE, entropy = TRUE)
  p <- as.numeric(predict(fit, Xva, type = "raw"))
  as.numeric(pROC::auc(yva, p, levels = c("Stay", "Cancel"),
                        direction = "<", quiet = TRUE))
}

cv_auc_nn <- function(size, decay) {
  mean(unlist(parallel::parLapply(nn_cl, nn_folds, fit_one_nn_fold,
                                   size = size, decay = decay)))
}

nn_grid <- expand.grid(size = c(5, 10, 20), decay = c(0.001, 0.01, 0.1))
nn_grid$cv_auc <- mapply(cv_auc_nn, nn_grid$size, nn_grid$decay)

parallel::stopCluster(nn_cl)
```

```{r nnet-cv-heatmap, fig.cap = "Figure 22: Cross-validation AUC surface for the neural network. The interior optimum confirms the architecture grid is adequate.", fig.width = 7, fig.height = 4}
kable(nn_grid, digits = 4,
      caption = "Table 5: Neural network — 5-fold CV AUC by (hidden units, weight decay).") %>%
  kable_styling(full_width = FALSE, font_size = 11)

ggplot(nn_grid, aes(x = factor(decay), y = factor(size), fill = cv_auc)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.3f", cv_auc)), colour = "white", size = 4) +
  scale_fill_viridis_c(option = "D",
                       limits = c(min(nn_grid$cv_auc), max(nn_grid$cv_auc))) +
  labs(title = "Neural Network — Cross-Validated AUC Surface",
       subtitle = "5-fold stratified CV on the full training set",
       x = "Weight Decay", y = "Hidden Units", fill = "CV AUC") +
  theme_minimal(base_size = 11)
```

## Final Model and Test Set Evaluation

```{r nnet-final-fit, cache = TRUE}
best_nn <- nn_grid[which.max(nn_grid$cv_auc), ]
set.seed(2026)
fit_nn <- nnet(x = nn_train$X, y = (nn_train$y == "Cancel") * 1L,
               size = best_nn$size, decay = best_nn$decay,
               maxit = 300, trace = FALSE, MaxNWts = 5000,
               linout = FALSE, entropy = TRUE)
cat(sprintf("Architecture: %d hidden units, decay = %.3f, weights = %d\n",
            best_nn$size, best_nn$decay, length(fit_nn$wts)))
```

```{r nnet-test}
prob_nn <- as.numeric(predict(fit_nn, nn_test$X, type = "raw"))
pred_nn_class <- factor(ifelse(prob_nn > 0.5, "Cancel", "Stay"),
                        levels = c("Stay", "Cancel"))
cm_nn  <- caret::confusionMatrix(pred_nn_class, nn_test$y, positive = "Cancel")
roc_nn <- pROC::roc(nn_test$y, prob_nn, levels = c("Stay", "Cancel"),
                    direction = "<", quiet = TRUE)
prc_nn <- PRROC::pr.curve(scores.class0 = prob_nn[nn_test$y == "Cancel"],
                          scores.class1 = prob_nn[nn_test$y == "Stay"],
                          curve = FALSE)

nn_results <- data.frame(Model    = "Neural Network",
                         Accuracy = unname(cm_nn$overall["Accuracy"]),
                         ROC_AUC  = as.numeric(roc_nn$auc),
                         F1       = unname(cm_nn$byClass["F1"]),
                         PR_AUC   = prc_nn$auc.integral)

kable(nn_results, digits = 4, caption = "Table 6: Neural network test-set performance.") %>%
  kable_styling(full_width = FALSE, font_size = 11)

cm_nn
```

## Permutation Variable Importance

A neural network's internal weights do not map directly to interpretable coefficients. To uncover which predictors the network relies on most heavily, we use permutation variable importance: for each predictor in turn, we randomly shuffle its values in the test set, recompute predictions, and measure the resulting drop in ROC-AUC. Predictors the network uses more than the GLM does are candidates for features whose contribution enters cancellation risk nonlinearly or interactively.

```{r nnet-permimp, cache = TRUE, fig.cap = "Figure 23: Permutation variable importance for the neural network. Red bars mark predictors the network relies on substantially more than the GLM — candidates for nonlinear or interactive use.", fig.width = 8, fig.height = 5}
set.seed(2026)
baseline_auc <- as.numeric(roc_nn$auc)

permute_auc <- function(j) {
  Xp <- nn_test$X; Xp[, j] <- sample(Xp[, j])
  p <- as.numeric(predict(fit_nn, Xp, type = "raw"))
  baseline_auc - as.numeric(pROC::auc(nn_test$y, p, levels = c("Stay", "Cancel"),
                                       direction = "<", quiet = TRUE))
}

perm_drops <- sapply(seq_len(ncol(nn_test$X)), permute_auc)
names(perm_drops) <- colnames(nn_test$X)

predictor_root <- function(col) {
  for (var in c("DepositType", "MarketSegment", "CustomerType", "arrival_season")) {
    if (startsWith(col, var)) return(var)
  }
  col
}

perm_df <- data.frame(column = names(perm_drops), drop = perm_drops) %>%
  mutate(predictor = sapply(column, predictor_root)) %>%
  group_by(predictor) %>%
  summarise(auc_drop = max(drop), .groups = "drop") %>%
  arrange(desc(auc_drop)) %>%
  mutate(predictor_clean = dplyr::recode(predictor,
    "LeadTime"               = "Lead Time",
    "DepositType"            = "Deposit Type",
    "MarketSegment"          = "Market Segment",
    "PreviousCancellations"  = "Previous Cancellations",
    "TotalOfSpecialRequests" = "Special Requests",
    "IsRepeatedGuest"        = "Repeated Guest",
    "BookingChanges"         = "Booking Changes",
    "DaysInWaitingList"      = "Days on Waiting List",
    "CustomerType"           = "Customer Type",
    "is_via_agent"           = "Booked via Agent",
    "arrival_season"         = "Arrival Season",
    "avg_temp_c"             = "Avg. Temperature",
    "precipitation_mm"       = "Precipitation"
  ))

glm_z <- summary(model_binomial)$coefficients[, "z value"]
glm_imp <- data.frame(column = names(glm_z), z = abs(glm_z)) %>%
  filter(column != "(Intercept)") %>%
  mutate(predictor = sapply(column, predictor_root)) %>%
  group_by(predictor) %>%
  summarise(glm_z = max(z), .groups = "drop")

perm_df <- perm_df %>%
  left_join(glm_imp, by = "predictor") %>%
  mutate(nn_rank   = rank(-auc_drop, ties.method = "first"),
         glm_rank  = rank(-glm_z,    ties.method = "first"),
         nn_higher = (glm_rank - nn_rank) >= 2)

ggplot(perm_df, aes(x = auc_drop, y = reorder(predictor_clean, auc_drop),
                    fill = nn_higher)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("TRUE" = "tomato", "FALSE" = "steelblue"),
                    labels = c("TRUE" = "Used more by NN than GLM",
                               "FALSE" = "Similar reliance"), name = NULL) +
  labs(title = "Neural Network — Permutation Variable Importance",
       subtitle = "Drop in test-set AUC when predictor is randomly shuffled",
       x = "AUC Drop", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

Predictors highlighted in red are those the network draws on substantially more than the Binomial GLM. These are the candidates for features whose contribution to cancellation risk is context-dependent: the same lead time means different things in different segments; the same booking-change pattern carries different implications for different customer types. For the hotel, this analysis identifies which variables would benefit from explicit interaction terms, if a more interpretable interactive model were desired in the future.

## Combined ROC Comparison

```{r nnet-roc-all, fig.cap = "Figure 24: ROC curves for all four cancellation models on the same held-out test set.", fig.width = 7, fig.height = 5}
plot(roc_bin, col = "steelblue",  lwd = 2,
     main = "ROC Curves: All Four Cancellation Models")
lines(roc_gam, col = "tomato",    lwd = 2)
lines(roc_rbf, col = "darkgreen", lwd = 2)
lines(roc_nn,  col = "purple",    lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright", bty = "n",
       legend = c(paste0("Binomial GLM   (AUC = ", round(auc(roc_bin), 3), ")"),
                  paste0("GAM            (AUC = ", round(auc(roc_gam), 3), ")"),
                  paste0("SVM (RBF)      (AUC = ", round(auc(roc_rbf), 3), ")"),
                  paste0("Neural Network (AUC = ", round(auc(roc_nn),  3), ")")),
       col = c("steelblue", "tomato", "darkgreen", "purple"), lwd = 2)
```

## Business Takeaway

The neural network achieves a test-set AUC of `r round(auc(roc_nn), 3)` — comparable to the RBF SVM and meaningfully above the additive models. Together, Models 5 and 6 confirm that a portion of cancellation risk is genuinely interactive in the booking features, and additive models leave that signal on the table. For a hotel deploying a cancellation risk system, the recommended architecture is two-tier: a flexible model (NN or RBF SVM) for the scoring engine, and the GLM for narrative explanations to commercial stakeholders. This captures the strengths of both model families rather than forcing a choice between accuracy and transparency.

---

# Model Comparison

All four cancellation models were fitted on the same training set, evaluated on the same held-out test set, and supplied with the same predictor set. Any performance difference reflects the model class, not the data each model had access to.

```{r model-comparison-table}
eval_glm_gam <- function(prob, label) {
  prob  <- as.numeric(prob)
  truth <- as.integer(test_bin$IsCanceled)
  ok    <- is.finite(prob) & !is.na(truth)
  prob  <- prob[ok]; truth <- truth[ok]
  pred  <- factor(ifelse(prob > 0.5, 1, 0), levels = c(0, 1))
  cm    <- caret::confusionMatrix(pred, factor(truth, levels = c(0, 1)), positive = "1")
  rocp  <- pROC::roc(truth, prob, quiet = TRUE)
  prc   <- PRROC::pr.curve(scores.class0 = prob[truth == 1],
                            scores.class1 = prob[truth == 0], curve = FALSE)
  data.frame(Model    = label,
             Accuracy = unname(cm$overall["Accuracy"]),
             ROC_AUC  = as.numeric(rocp$auc),
             F1       = unname(cm$byClass["F1"]),
             PR_AUC   = prc$auc.integral)
}

comparison_table <- rbind(
  eval_glm_gam(pred_probs_bin, "Binomial GLM"),
  eval_glm_gam(gam_probs,      "GAM"),
  svm_results,
  nn_results
)
rownames(comparison_table) <- NULL

kable(comparison_table, digits = 4,
      caption = "Table 7: Test-set performance of all four cancellation models on identical held-out data. Higher is better on every metric. The best-performing row is highlighted.") %>%
  kable_styling(full_width = FALSE, font_size = 11) %>%
  row_spec(which.max(comparison_table$ROC_AUC), bold = TRUE, background = "#EFF6FF")
```

## Decomposing the Performance Gains

Three patterns emerge from Table 7 that directly validate the ladder of assumptions in Table 0.

**The largest single jump comes from relaxing linearity.** The GAM's AUC gain over the GLM of `r round(comparison_table$ROC_AUC[2] - comparison_table$ROC_AUC[1], 3)` points comes from replacing three linear continuous terms with smooth functions — no change to the additive structure of the model. A non-trivial share of the cancellation signal is in the *shape* of continuous relationships, particularly the lead-time plateau that Figure 17 quantified precisely.

**The second jump comes from allowing interactions.** The RBF SVM and Neural Network both exceed the GAM, confirming that cancellation risk depends on combinations of predictors that no additive model can capture. The fact that two architecturally different models converge on similar AUCs — `r round(comparison_table$ROC_AUC[4], 3)` and `r round(comparison_table$ROC_AUC[5], 3)` — suggests we are observing the empirical ceiling of what this predictor set supports, not an artefact of either model class.

**The linear-kernel SVM serves as a deliberate control.** Its AUC of `r round(comparison_table$ROC_AUC[3], 3)` sits close to the Binomial GLM's `r round(comparison_table$ROC_AUC[1], 3)`, despite using a completely different training objective. This rules out the interpretation that the RBF SVM's gains are due to margin-maximisation rather than kernel nonlinearity.

## Accuracy vs. Interpretability

```{r comparison-tradeoff-table, echo = FALSE}
tradeoff <- data.frame(
  Model      = c("Binomial GLM", "GAM", "SVM (RBF)", "Neural Network"),
  AUC        = round(comparison_table$ROC_AUC[comparison_table$Model %in%
                 c("Binomial GLM", "GAM", "SVM (RBF)", "Neural Network")], 3),
  Interpret  = c("Full — odds ratios", "Partial — smooth plots + odds ratios",
                 "None — permutation importance only", "None — permutation importance only"),
  Retrain    = c("Seconds", "Seconds", "Hours (subsample required)", "Minutes"),
  Cadence    = c("Daily", "Daily", "Weekly at most", "Daily")
)

kable(tradeoff,
      caption = "Table 8: Accuracy vs. interpretability trade-off across the four cancellation models.",
      col.names = c("Model", "Test AUC", "Interpretability", "Retraining Cost",
                    "Feasible Cadence")) %>%
  kable_styling(full_width = TRUE, font_size = 11)
```

The accuracy ranking and the operational ranking are not the same. The most accurate model is the slowest to retrain and the hardest to explain. A two-tier deployment — flexible model (RBF SVM or Neural Network) for scoring decisions, Binomial GLM for narrative explanations — captures the strengths of both without forcing a choice. If only one model can be maintained, the GAM is the defensible single-model compromise: most of the accuracy gain at a fraction of the interpretability cost.

---

# Conclusion

This report applied six models to a single city hotel dataset, answering three operationally distinct questions across the full booking lifecycle.

**On price (Model 1).** The Linear Model explains `r round(summary(model_lm)$adj.r.squared * 100, 1)`% of room price variance, with an out-of-sample prediction error of approximately ±€`r round(abs(rmse_eur), 1)` per night. Market segment is the dominant price driver — Direct bookings sustain the highest rates. A negative lead-time effect, robust to full multivariate control, supports structured early-bird pricing tiers: last-minute bookings command a modest premium, and this is defensible on the evidence.

**On stay length (Model 2).** The quasi-Poisson GLM — necessary because the data exhibit severe overdispersion (dispersion ratio `r round(disp_ratio, 1)`) — confirms that Summer leisure segments stay longest and Corporate guests shortest. Meal plan and customer type are stronger predictors than weather or geography. The model predicts stay length to within `r round(mae_pois, 1)` nights on unseen data. The operational implication: long-stay promotions should target Summer Online TA leisure guests, not Corporate travellers whose short stays are structural.

**On cancellation (Models 3–6).** The four-model progression is the report's core analytical contribution. The Binomial GLM identifies *which* predictors drive cancellation and by how much (AUC `r round(auc(roc_bin), 3)`). The GAM shows *how* they drive it — revealing a genuine nonlinear plateau in the lead-time relationship and a non-monotone ADR effect invisible to linear models (AUC `r round(auc(roc_gam), 3)`). The SVM and Neural Network show *how much additional accuracy* is available when feature interactions are modelled — both reach AUCs of approximately `r round(max(auc(roc_rbf), auc(roc_nn)), 3)`, confirming that cancellation risk is genuinely interactive in the booking features, not merely nonlinear in individual predictors. The performance gain decomposes cleanly: roughly half comes from allowing nonlinear continuous effects (the GAM step) and half from allowing interactions (the SVM and NN step).

**A unifying observation.** The same predictor set — lead time, market segment, deposit type, special requests, previous cancellations — appears as significant across all three response variables. Price, stay length, and cancellation risk are not three independent phenomena: they are facets of a single underlying booking-quality dimension. A guest who books early, books directly, submits special requests, and has no prior cancellation history pays more, stays longer, and is dramatically less likely to cancel. Revenue management that optimises for any one of these dimensions in isolation is leaving value on the table.

**Operational recommendation.** For real-time cancellation scoring, the RBF SVM or Neural Network should serve as the scoring engine. For communication with commercial stakeholders, the Binomial GLM's odds ratios are the language to use. For the hotel that can maintain only one model, the GAM is the best single-model compromise. For pricing and stay-length decisions, the Linear Model and quasi-Poisson GLM provide directly actionable coefficient estimates that require no further statistical translation.

---

# Limitations

**Weather granularity.** Weather variables are monthly aggregates: every booking in the same arrival month receives identical weather values. This is sufficient to capture seasonal climate effects but provides no within-month variation. The marginal contribution of weather to the cancellation models should be read with this constraint in mind. `sunshine_hours`, the third weather variable, is not used in any model because in Lisbon's climate it is highly correlated with `avg_temp_c` — including both would introduce multicollinearity without adding information.

**The Non Refund recording artefact.** All four cancellation models exploit the near-perfect cancellation rate of Non Refund deposits. This inflates absolute AUC values by an unknown amount. Because all four models are exposed equally, their relative ranking is unaffected — but a hotel deploying these models after changing its deposit recording convention should expect lower absolute AUCs than reported in Table 7.

**SVM subsampling.** The kernel SVM was tuned and fitted on a stratified 5,000-row subsample due to computational constraints. The reported test-set performance is directly comparable to the other models (same test set), but is a lower-variance estimate of full-data SVM performance rather than the exact figure a full-data fit would produce.

**Single-property scope.** The dataset covers one Lisbon city hotel over 25 months. Findings about segments, seasons, and weather sensitivity are specific to this property and period. Extending these models to other properties or to post-2020 booking behaviour would require revalidation, not just retraining.

---

# AI Usage Statement

Generative AI tools — primarily Claude (Anthropic) — were used throughout this project in the following ways.

**Code assistance.** AI was used to debug R code, suggest idiomatic tidyverse patterns, and troubleshoot chunk caching issues in R Markdown. All generated code was reviewed, tested, and modified by the team before inclusion. The quasi-Poisson correction, the `drop1()` test structure, and the GAM smooth extraction approach were all verified against course material and primary R documentation (`?mgcv::gam`, `?glm`).

**Prose drafting.** AI was used to draft initial versions of section introductions and plot interpretation paragraphs. All prose was substantially rewritten by the section lead to reflect the team's analytical judgement and the actual numbers produced by the models. AI-generated interpretations were treated as starting points, not final text.

**Concept clarification.** AI was used to clarify the distinction between overdispersion and zero-inflation, the interpretation of EDF in GAMs, and the relationship between SVM decision values and probability calibration.

**What was not done with AI.** All modelling decisions — predictor selection, reference level choices, the decision to use quasi-Poisson, the four-model cancellation ladder structure — were made by the team based on course material and data examination. The central analytical argument (that the performance progression decomposes into a nonlinearity gain and an interaction gain) was developed by the team, not suggested by AI.

---

# References

Antonio, N., de Almeida, A., & Nunes, L. (2019). Hotel booking demand datasets. *Data in Brief*, 22, 41–49. https://doi.org/10.1016/j.dib.2018.11.126

Open-Meteo Historical Weather API. Retrieved from https://open-meteo.com

Vapnik, V. (1995). *The Nature of Statistical Learning Theory*. Springer.

Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R* (2nd ed.). Chapman and Hall/CRC.
